---
title: "Accessibility"
description: >
  What every plot gets for free (a labelled, described SVG, a tagged PDF, a
  generated text alternative, and a keyboard- and screen-reader-navigable
  interactive widget), what that baseline does not cover, and how to author the
  rest.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A chart that a screen-reader user cannot perceive, or that a keyboard user
cannot operate, excludes people. The vellum ecosystem treats accessibility as
part of the output contract rather than an afterthought: **every plot you compile
starts from a usable baseline** — a title and a generated text alternative, an SVG
that announces itself as an image, a tagged PDF, and a widget you can operate
without a mouse.

A baseline is not compliance. A generated description can say what is plotted but
not what it means, and nothing here checks your colour contrast or decides whether
your takeaway is legible. What the ecosystem removes is the boilerplate excuse:
the mechanics are already in place, so the remaining work is editorial. This
article is the cross-package guide to both halves.

The work is split across the three packages along the same seam as everything
else:

- **vellumplot** (this package) is the *author*. It generates the plot's text
  alternative and hands it, with the title, to the scene.
- **vellum** is the *emitter*. It turns that title/description into an accessible
  SVG and a tagged PDF.
- **vellumwidget** is the *interactive layer*. It makes the widget keyboard- and
  screen-reader-navigable.

## Alt text: the text alternative

The single most important thing a chart needs is a **text alternative**, a
prose sentence or two describing what it shows (WCAG 1.1.1). vellumplot writes one
for every plot automatically, from the spec it already has:

```{r}
p <- vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp)

plot_alt(p)
```

The generated description names the chart type, what is mapped to each axis and
to colour/size, the number of observations, and any faceting. It is a *sensible
default*, not a substitute for editorial judgement. The auto text says *what* is
plotted, not *what it means*. When you know the takeaway, write it:

```{r}
p <- p |>
  labs(
    title = "Heavier cars use more fuel",
    alt = paste(
      "Scatter plot of fuel economy against weight for 32 cars.",
      "Fuel economy falls roughly linearly as weight rises, from about",
      "34 mpg at 1.5 tons to about 10 mpg at 5 tons."
    )
  )

plot_alt(p)
```

`labs(alt =)` always wins over the automatic text. `plot_alt()` returns whichever
applies, so you can inspect exactly what assistive technology will receive.

## What the accessible SVG emits

At the compile seam the plot **title** becomes the scene's accessible *name* and
`plot_alt()` becomes its *description*. vellum emits these as the reliable
screen-reader pattern: `role="img"` with `<title>` and `<desc>` referenced by
`aria-labelledby`:

```{r}
f <- tempfile(fileext = ".svg")
render_plot(p, f)

svg <- paste(readLines(f), collapse = "\n")
# the opening <svg> tag plus its <title>/<desc>
regmatches(svg, regexpr("<svg[^>]*>.*?</desc>", svg, perl = TRUE))
```

A screen reader announces the title as the image's name and the description on
request. This happens for every compiled plot; you never have to remember to
turn it on. (A plot with no title still carries the auto-generated description,
so it is never a mute image.)

## Accessible PDF

The same title/description drive a **tagged PDF** (the basis of PDF/UA) when you
export to `.pdf`. The chart becomes a `Figure` in the document's structure tree,
carrying the description as its `Alt` text, so a PDF reader can announce it:

```{r}
#| eval: false
render_plot(p, "fuel.pdf") # a tagged PDF: /StructTreeRoot + a Figure with Alt
```

Tagging is automatic and additive. A plot with no title/description exports as
an ordinary, untagged PDF exactly as before. (Strict PDF/UA-1 *validation* is a
planned follow-up; the tag tree and `Alt` ship today.)

## The interactive widget

A static image with alt text is enough for a picture, but an *interactive* chart
must also be operable and explorable without a mouse. `vellumwidget::as_widget()`
(accessibility on by default, `a11y = TRUE`) makes the widget a first-class
interactive chart:

```{r}
#| eval: false
library(vellumwidget)

vplot(mtcars) |>
  mark_point(x = wt, y = mpg, tooltip = rownames(mtcars), data_id = rownames(mtcars)) |>
  labs(title = "Fuel economy", alt = "Fuel economy against weight for 32 cars.") |>
  as_widget()
```

That widget:

- announces itself as an **interactive chart**
  (`role="graphics-document"`), labelled from the title/description vellumplot
  already emitted (or an explicit `as_widget(alt =)`);
- makes **every mark focusable**. Tab into the chart, then use the **arrow
  keys** to move between marks; each is a `graphics-symbol` labelled with its
  tooltip. **Enter** or **Space** selects the focused mark; **Escape** leaves
  traversal mode;
- announces the focused and selected mark through a polite **`aria-live`**
  region, so a screen-reader user hears what they land on;
- ships a visually-hidden **data table** listing every mark, so the underlying
  data is available even without traversing the chart.

Setting `a11y = FALSE` restores the previous behaviour exactly, but there is
rarely a reason to.

## Authoring checklist

- **Give plots a `title`.** It becomes the accessible name.
- **Write a real `alt`** when the automatic one misses the point: say what the
  chart *shows*, not only what it plots. Inspect it with `plot_alt()`.
- **Set `tooltip` / `data_id`** on interactive marks: the tooltip is what a
  screen-reader user hears when focusing a mark, and appears in the data table.
- **Don't rely on colour alone.** Pair a colour encoding with shape, or label
  the extremes, so the chart is legible without colour perception.

## See also

- vellum's *The scene contract* vignette: the SVG/PDF a11y attributes and the
  additivity invariant, from the emitter's side.
- vellumwidget's interactivity article: the widget's full keyboard and screen-reader
  model.
