---
title: "Accessibility"
description: >
  What every plot gets for free (a labelled, described SVG, a tagged PDF, a
  generated text alternative, and a keyboard- and screen-reader-navigable
  interactive widget), what that baseline does not cover, and how to author the
  rest.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A chart that a screen-reader user cannot perceive, or that a keyboard user
cannot operate, excludes people. The vellum ecosystem treats accessibility as
part of the output contract rather than an afterthought: **every plot you compile
starts from a usable baseline** — a title and a generated text alternative, an SVG
that announces itself as an image, a tagged PDF, and a widget you can operate
without a mouse.

A baseline is not compliance. A generated description can say what is plotted but
not what it means. `plot_lint()` now checks your colour contrast and text
legibility (see below), but nothing decides whether your *takeaway* is legible.
What the ecosystem removes is the boilerplate excuse: the mechanics are already
in place, so the remaining work is editorial. This article is the cross-package
guide to both halves.

The work is split across the three packages along the same seam as everything
else:

- **vellumplot** (this package) is the *author*. It generates the plot's text
  alternative and hands it, with the title, to the scene.
- **vellum** is the *emitter*. It turns that title/description into an accessible
  SVG and a tagged PDF.
- **vellumwidget** is the *interactive layer*. It makes the widget keyboard- and
  screen-reader-navigable.

## Alt text: the text alternative

The single most important thing a chart needs is a **text alternative**, a
prose sentence or two describing what it shows (WCAG 1.1.1). vellumplot writes one
for every plot automatically, from the spec it already has:

```{r}
p <- vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp)

plot_alt(p)
```

The generated description names the chart type, what is mapped to each axis and
to colour/size, the number of observations, and any faceting. It is a *sensible
default*, not a substitute for editorial judgement. The auto text says *what* is
plotted, not *what it means*. When you know the takeaway, write it:

```{r}
p <- p |>
  labs(
    title = "Heavier cars use more fuel",
    alt = paste(
      "Scatter plot of fuel economy against weight for 32 cars.",
      "Fuel economy falls roughly linearly as weight rises, from about",
      "34 mpg at 1.5 tons to about 10 mpg at 5 tons."
    )
  )

plot_alt(p)
```

`labs(alt =)` always wins over the automatic text. `plot_alt()` returns whichever
applies, so you can inspect exactly what assistive technology will receive.

## What the accessible SVG emits

At the compile seam the plot **title** becomes the scene's accessible *name* and
`plot_alt()` becomes its *description*. vellum emits these as the reliable
screen-reader pattern: `role="img"` with `<title>` and `<desc>` referenced by
`aria-labelledby`:

```{r}
f <- tempfile(fileext = ".svg")
render_plot(p, f)

svg <- paste(readLines(f), collapse = "\n")
# the opening <svg> tag plus its <title>/<desc>
regmatches(svg, regexpr("<svg[^>]*>.*?</desc>", svg, perl = TRUE))
```

A screen reader announces the title as the image's name and the description on
request. This happens for every compiled plot; you never have to remember to
turn it on. (A plot with no title still carries the auto-generated description,
so it is never a mute image.)

## Accessible PDF

The same title/description drive a **tagged PDF** (the basis of PDF/UA) when you
export to `.pdf`. The chart becomes a `Figure` in the document's structure tree,
carrying the description as its `Alt` text, so a PDF reader can announce it:

```{r}
#| eval: false
render_plot(p, "fuel.pdf") # a tagged PDF: /StructTreeRoot + a Figure with Alt
```

Tagging is automatic and additive. A plot with no title/description exports as
an ordinary, untagged PDF exactly as before. (Strict PDF/UA-1 *validation* is a
planned follow-up; the tag tree and `Alt` ship today.)

## Colour, contrast, and legibility

Alt text and tags make a plot *perceivable to assistive technology*. A separate
question is whether the picture itself is legible — text large enough to read,
contrast high enough to see, and information that does not vanish for a
colour-blind viewer. Three tools form a loop: **see** the problem, **flag** it,
**fix** it.

### See it: `render_plot(cvd = )`

About 1 in 12 men has some form of colour-vision deficiency, most commonly
red–green. `render_plot(cvd = )` renders a `.png` through a simulation of how a
palette reads for such a viewer — `"protanopia"`, `"deuteranopia"`,
`"tritanopia"`, or `"achromatopsia"` — so you can see a failing palette instead
of guessing:

```{r}
#| eval: false
p <- vplot(mtcars) |> mark_point(x = wt, y = mpg, color = factor(cyl))

render_plot(p, "normal.png")
render_plot(p, "deuteranopia.png", cvd = "deuteranopia") # red-green preview
```

Simulation is raster-only (it is a pixel operation), so it is ignored for
`.svg`/`.pdf`.

### Flag it: `plot_lint()`

`plot_lint()` compiles the plot and reports the legibility problems a green test
suite hides: text below a readable size, contrast under the WCAG threshold, two
palette colours a colour-blind reader cannot tell apart, labels that overlap or
fall off the panel (all judged in real device pixels by the engine), plus
grammar-level mistakes such as an encoding with a single level or a legend too
long to read.

```{r}
p_bad <- vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(rep("all cars", nrow(mtcars)))) |>
  theme(axis.text = element_text(size = 4))

plot_lint(p_bad)
```

A well-made plot lints clean:

```{r}
plot_lint(vplot(mtcars) |> mark_point(x = wt, y = mpg))
```

Each finding is a row: `rule`, `severity`, `node`, `message`, and the device-px
box the finding refers to. `min_text_px` and `min_contrast` set the thresholds,
and every other argument of `vellum::vl_lint()` comes through too — `exclude` to
accept a finding you have already judged, `cvd` to choose which colour-vision
deficiency to check, `rules` to run one rule on its own.

Because the findings carry their boxes, the report can be drawn onto the plot
instead of read:

```{r}
#| fig.alt: "The plot above with boxes drawn around each linted element."
scene <- vellum::as_vellum_scene(p_bad)
vellum::display(vellum::vl_lint_overlay(scene, plot_lint(p_bad)))
```

And it can gate a test suite, which is the point of flagging things at all:

```{r}
#| eval: false
test_that("the figure is legible", {
  vellum::vl_lint_assert(my_plot())
})
```

The grammar rules are not a separate mechanism. vellumplot registers them in the
engine's own rule registry, so `vellum::vl_lint_rules()` lists them, `rules =`
selects them, and a plain `vellum::vl_lint()` of a compiled plot reports the
encoding problems alongside the geometric ones:

```{r}
subset(vellum::vl_lint_rules(), tags == "grammar", c("rule", "description"))
```

### Fix it: a redundant, non-colour encoding

The robust fix for a colour encoding that fails under CVD or in greyscale is to
carry the same information a second way. Map the `pattern` aesthetic alongside
`fill`, and reach for `pattern_hatch()` — a crisp **vector** hatch that stays
sharp in PDF and survives being printed in black and white (unlike the raster
tile patterns):

```{r}
d <- data.frame(grp = c("A", "B", "C"), n = c(8, 5, 11))

vplot(d) |>
  mark_bar(x = grp, y = n, fill = grp, pattern = grp) |>
  scale_pattern(values = list(
    pattern_hatch(angle = 0),
    pattern_hatch(angle = 45),
    pattern_hatch(angle = 90)
  ))
```

Now each bar is distinguished by hatch *and* colour: the plot survives the CVD
preview above, and `plot_lint()`'s contrast rule has less to complain about.

## The interactive widget

A static image with alt text is enough for a picture, but an *interactive* chart
must also be operable and explorable without a mouse. `vellumwidget::as_widget()`
(accessibility on by default, `a11y = TRUE`) makes the widget a first-class
interactive chart:

```{r}
#| eval: false
library(vellumwidget)

vplot(mtcars) |>
  mark_point(x = wt, y = mpg, tooltip = rownames(mtcars), data_id = rownames(mtcars)) |>
  labs(title = "Fuel economy", alt = "Fuel economy against weight for 32 cars.") |>
  as_widget()
```

That widget:

- announces itself as an **interactive chart**
  (`role="graphics-document"`), labelled from the title/description vellumplot
  already emitted (or an explicit `as_widget(alt =)`);
- makes **every mark focusable**. Tab into the chart, then use the **arrow
  keys** to move between marks; each is a `graphics-symbol` labelled with its
  tooltip. **Enter** or **Space** selects the focused mark; **Escape** leaves
  traversal mode;
- announces the focused and selected mark through a polite **`aria-live`**
  region, so a screen-reader user hears what they land on;
- ships a visually-hidden **data table** listing every mark, so the underlying
  data is available even without traversing the chart.

Setting `a11y = FALSE` restores the previous behaviour exactly, but there is
rarely a reason to.

## Authoring checklist

- **Give plots a `title`.** It becomes the accessible name.
- **Write a real `alt`** when the automatic one misses the point: say what the
  chart *shows*, not only what it plots. Inspect it with `plot_alt()`.
- **Set `tooltip` / `data_id`** on interactive marks: the tooltip is what a
  screen-reader user hears when focusing a mark, and appears in the data table.
- **Don't rely on colour alone.** Pair a colour encoding with a redundant
  non-colour one — `shape`, or a `pattern` mapped through `scale_pattern()` with
  `pattern_hatch()` — or label the extremes, so the chart is legible without
  colour perception. Preview it with `render_plot(cvd = )`.
- **Lint before you ship.** Run `plot_lint()` to catch tiny text, low contrast,
  overlaps and dead encodings while they are still cheap to fix.

## See also

- vellum's *The scene contract* vignette: the SVG/PDF a11y attributes and the
  additivity invariant, from the emitter's side.
- vellumwidget's interactivity article: the widget's full keyboard and screen-reader
  model.
