---
title: "Animation"
description: >
  Non-reactive keyframe animation: compile a plot into one keyframe per state,
  train the scales once over every state and freeze them, then tween and encode
  the in-between frames (GIF / APNG) in vellum's parallel Rust backend.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

An animated plot is a sequence of still plots played in quick succession. In
vellumplot the stills are **keyframes** — one compiled plot per *state* — and the
frames between them are **tweened**: their marks are interpolated so a point
glides from where it sits in one state to where it sits in the next.

This is *non-reactive* keyframe animation. The states are decided up front (the
levels of a column), and the scales are trained **once over every state and then
frozen**. Nothing retrains between frames: the axis breaks, the colour ramp and
the size legend are the same in every frame, so the eye can read motion against a
fixed backdrop. Animation is not a live/reactive runtime — it is a fixed set of
states compiled ahead of time.

## The pipeline

Four verbs, added to any plot:

* `transition_states(states)` — marks the column whose levels are the states.
* `ease_aes(ease)` — chooses the easing (how the interpolation accelerates).
* `animate(nframes, fps)` — compiles the keyframes and returns a
  `vellum_animation`.
* `anim_save(file, animation)` — tweens and encodes the frames to a file.

```{r}
#| eval: false
library(gapminder)

vplot(gapminder) |>
  mark_point(x = gdpPercap, y = lifeExp, size = pop, color = continent) |>
  scale_x_continuous(transform = "log10") |>
  transition_states(year, transition_length = 2, state_length = 1) |>
  ease_aes("cubic-in-out") |>
  animate(nframes = 100, fps = 25) |>
  (\(a) anim_save("gapminder.gif", a))()
```

The output format follows the file extension: `.gif` for a looping animated GIF
(universally viewable) or `.png` for an animated PNG (APNG — lossless, with
transparency). The tween, the parallel render across CPU cores, and the streaming
encode all happen in one pass in vellum's Rust backend.

## What a keyframe looks like

`animate()` compiles one keyframe per state. Each is an ordinary scene, so you can
inspect one directly — here the first state of an `mtcars`-by-cylinder animation:

```{r}
#| fig-alt: "A scatter of weight against miles-per-gallon, one keyframe of an animation."
a <- vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = hp, color = factor(cyl)) |>
  transition_states(cyl) |>
  ease_aes("cubic-in-out") |>
  animate(nframes = 60, fps = 20)

# a@scenes holds one vellum scene per state; render the first as a still.
vellum::render(a@scenes[[1]], tempfile(fileext = ".png"))
```

## Frozen scales — the non-reactive guardrail

Because the scales are trained once over the pooled states, the domain is the
*union* across every state. A point that is extreme in one state and typical in
another keeps a stable position reference throughout: the axes never jump. This is
the observable proof that the animation stayed non-reactive — compare the axis
breaks of any two keyframes and they are identical, even though the two states'
data cover very different ranges.

Per-state *statistics* still recompute (a smooth is refit per state, a bin count
re-tallied), but no scale *domain* retrains. Retraining a scale in response to the
data of the moment would be a reactive runtime, which vellumplot is deliberately
not.

## Easing

`ease_aes()` shapes how the interpolation fraction moves through each transition.
`"linear"` is a constant rate; a family name with a direction suffix eases in, out,
or both:

* families: `quad`, `cubic`, `quart`, `quint`, `sine`, `expo`, `circ`, `back`,
  `elastic`, `bounce`
* directions: `-in`, `-out`, `-in-out`

`"cubic-in-out"` — slow start, quick middle, slow finish — reads as the most
natural motion for most plots.

## What tweens, and what does not

Between two keyframes, an element's **position**, **size**, **alpha**, and
**colour** interpolate — colour perceptually, in Oklab, so a hue transition stays
vivid instead of passing through grey. Discrete attributes (a marker shape, a line
type, a text label) do not interpolate continuously; they snap at the midpoint.

The clean case is a **stable set of elements** with the same identity in every
state (the same countries each year, the same cars in each cylinder group). When
the set of elements changes between states, or the plot is faceted, the transition
is not a smooth morph — that is outside the current scope.

## Timing

`transition_length` and `state_length` are *relative* durations: how long the
moving segments last versus the held pauses on each state. `nframes` is the total
frame count and `fps` the playback rate, so the animation runs for `nframes / fps`
seconds. `wrap = TRUE` (the default) loops the last state back to the first.
