---
title: "Effects and themes"
description: >
  The non-data look of a plot: built-in themes, custom theme elements, layer
  effects like glow and shadow, gradient fills, and the one-line hand-drawn
  sketch mode.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

Everything so far has been about mapping data. This article is about the rest:
the panel, the gridlines, the fonts, and a set of render effects that change how
marks are painted. vellumplot keeps these separate. A *theme* controls the
non-data furniture of the whole plot; a *layer effect* changes how one mark is
drawn; a *gradient* is a fancy fill value; and *sketch* mode turns the whole
thing hand-drawn.

## Built-in themes

A theme sets the look of everything that is not a mark. vellumplot ships several.
`theme_gray()` is the default (grey panel, white gridlines); `theme_minimal()`
drops the panel fill; `theme_bw()` is a white panel with light grey gridlines;
`theme_classic()` gives axis lines and no gridlines; `theme_void()` strips
everything but the marks, legend, and titles.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  theme_minimal()
```

## Customising with theme elements

`theme()` overrides individual slots, and each slot is described by a typed
*element*: `element_text()`, `element_line()`, `element_rect()`, or
`element_blank()` to draw nothing. Slot names follow the familiar dotted scheme
(`panel.grid.minor`, `axis.title`, `plot.title`, and so on), and any property
left `NULL` is inherited from its parent in the theme tree.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  theme_bw() |>
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(color = "grey30")
  ) |>
  labs(title = "Fuel economy")
```

## Layer effects

Effects change how a single mark is painted, and they are passed to a mark's
`effects` argument as a list. They apply to stroked and point marks.
`glow()` draws widened, low-opacity copies beneath the mark for a neon halo;
`shadow()` adds an offset drop shadow; `outline()` puts a contrasting halo
behind the mark so it stays legible over a busy backdrop. `motion()` and
`echo()` draw a fading trail of copies marching off along a direction — a
speed-blur (`motion()`, many close copies) or discrete ghost repeats (`echo()`).

```{r}
vplot(pressure) |>
  mark_line(
    x = temperature, y = pressure,
    effects = list(shadow(), outline(color = "white", size = 2))
  )
```

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = 4, effects = list(motion(x = 4)))
```

`glow()` pairs naturally with `theme_cyberpunk()`, a dark neon theme.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl), effects = list(glow())) |>
  theme_cyberpunk()
```

## Gradient fills

A gradient is an unscaled *value* for the `fill` aesthetic, not a mapped scale.
Pass `linear_gradient()` or `radial_gradient()` directly as a fill and the
region is painted with it as one paint. Using a transparent stop gives the
"fade out under a line" look.

```{r}
vplot(pressure) |>
  mark_area(
    x = temperature, y = pressure,
    fill = linear_gradient(c("#00e5ff", "#00e5ff00"))
  ) |>
  mark_line(x = temperature, y = pressure, color = "#00e5ff") |>
  theme_cyberpunk()
```

Because a gradient is one paint per region, it cannot be mapped to a data
column; for that you want a colour scale (see
**[Scales and guides](scales-and-guides.html)**).

## Pattern (hatch) fills

A *pattern* is the texture counterpart of a gradient: another unscaled `fill`
value, built by the `pattern_*()` family. Distinguishing regions by texture (not
only hue) keeps a plot legible in greyscale print and under colour-vision
deficiency. Pass one directly as a `fill`:

```{r}
bars <- data.frame(method = c("A", "B", "C"), score = c(4, 7, 5))
vplot(bars) |>
  mark_bar(x = method, y = score, fill = pattern_crosshatch(color = "grey20"))
```

`pattern_stripe()`, `pattern_crosshatch()`, `pattern_grid()`, `pattern_dot()`,
and `pattern_checker()` each take a `color`, a transparent-or-solid `bg`, and a
`spacing` (in mm by default); stripes/crosshatch also take an `angle` restricted
to `0`, `45`, `90`, or `135`. Patterns work on any filled mark — bars, areas,
ribbons, rects, tiles, boxplots, violins, ridgelines, half-eyes, hulls, and `sf`
polygons — and render on every backend including PDF. As with gradients, a
pattern is one fill per region (an unscaled value, not a mapped channel). For a
custom motif, build the tile yourself with `vl_pattern()`.

## Clipping and masking to a shape

`clip_to()` restricts a plot's marks to a geometry instead of the panel
rectangle — the way to pour a raster or tile heatmap into a region outline. Give
it an `sf` object or a data frame of `x`/`y` polygon vertices.

```{r}
grid <- expand.grid(x = 1:24, y = 1:24)
grid$z <- with(grid, sin(x / 4) + cos(y / 4))
diamond <- data.frame(x = c(12, 22, 12, 2), y = c(2, 12, 22, 12))

vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  clip_to(diamond)
```

`invert = TRUE` keeps the marks *outside* the shape (punching it out as a hole).
`set_mask()` instead applies a soft radial mask — a vignette that fades the panel
towards its edges, with `feather` setting how gentle the fade is:

```{r}
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  set_mask(feather = 0.5)
```

`clip_to()` and `set_mask()` mask the whole panel; `clip_layer()` masks only the
most-recently-added layer, so one raster layer can be clipped to a shape while
the axes and other layers stay full-bleed:

```{r}
set.seed(1)
pts <- data.frame(x = runif(60, 1, 24), y = runif(60, 1, 24))
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  clip_layer(diamond) |>
  mark_point(data = pts, x = x, y = y, size = 1.5, color = "white")
```

All three work under cartesian coordinates (a smooth feathered edge on an
arbitrary polygon is not available yet, so `clip_to()` / `clip_layer()` are
hard-edged; reach for `set_mask()` when you want a soft fade).

## Hand-drawn sketch mode

`theme_sketch()` turns an entire plot hand-drawn in one line: wobbly gridlines,
axis lines, and marks, on a warm paper background. The look is generated
natively in the engine, so it is exact and works across PNG, SVG, and PDF.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(cyl)) |>
  theme_sketch()
```

For finer control, `sketch()` is the vocabulary: `roughness`, `bowing`,
`fill_style` (`"hachure"`, `"crosshatch"`, `"zigzag"`, and more), and so on.
Pass a `sketch()` value to a single mark's `sketch =` argument to rough up just
that layer, or to an `element_line()` / `element_rect()` slot inside `theme()`.
Set `sketch = NA` on a mark to force it crisp even under a plot-wide
`theme_sketch()`.

```{r}
vplot(mtcars) |>
  mark_bar(
    x = factor(cyl), fill = factor(cyl),
    sketch = sketch(roughness = 2, fill_style = "crosshatch")
  )
```
