---
title: "Facets and composition"
description: >
  Two ways to make many panels: facet one plot across a variable with shared
  or free scales, and combine separate plots into one figure with concat,
  insets, and repeats.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("quill", quietly = TRUE)
)
library(quill)
```

There are two different reasons to end up with a grid of panels, and quill
keeps them separate. *Faceting* takes one plot and splits it by a variable, so
every panel shares the same encodings and (by default) the same scales.
*Composition* takes several independent plots, each its own spec, and arranges
them into a single figure. Use faceting for small multiples of the same view;
use composition for a dashboard of different views.

## Faceting

`facet_wrap()` lays panels out in a ribbon wrapped to `ncol` or `nrow`.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  facet_wrap(~cyl, ncol = 3)
```

`facet_grid()` puts panels on a 2-D grid defined by a `rows ~ cols` formula,
which is the right tool when you are crossing two variables.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  facet_grid(vs ~ am)
```

### Shared or free scales

By default position scales are shared across panels so the axes line up and
panels are comparable. When panels cover very different ranges, free them per
panel with `scales = "free_x"`, `"free_y"`, or `"free"`.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  facet_wrap(~cyl, scales = "free")
```

`scales = "free_y"` is sugar for the underlying `resolve_scale()` lattice,
which lets you set the policy per aesthetic directly. These two are equivalent:

```{r}
#| eval: false
vplot(mtcars) |> mark_point(x = wt, y = mpg) |>
  facet_wrap(~cyl, scales = "free_y")

vplot(mtcars) |> mark_point(x = wt, y = mpg) |>
  facet_wrap(~cyl) |> resolve_scale(y = "independent")
```

Colour and size scales (and their legends) stay shared across panels, so one
legend describes the whole figure.

## Composition

To place *different* plots side by side, build each one and combine them.
`concat()` (and its friends `hconcat()` and `vconcat()`) arrange plots into a
grid; by default they collect the legends into one shared guide.

```{r}
p1 <- vplot(mtcars) |> mark_point(x = wt, y = mpg, color = factor(cyl))
p2 <- vplot(mtcars) |> mark_boxplot(x = factor(cyl), y = mpg)
p3 <- vplot(mtcars) |> mark_histogram(x = mpg, bins = 10)

concat(p1, p2, p3, ncol = 3)
```

Control the shape with `ncol`/`nrow`, the relative sizes with
`widths`/`heights`, and irregular layouts with a `design` string.
`plot_spacer()` drops an empty cell, and `wrap_plots()` takes a list of plots
when you have assembled them programmatically.

```{r}
concat(p1, p2, ncol = 1, heights = c(2, 1))
```

### Insets

`inset()` floats one plot over another, positioned by fractional coordinates of
a reference box. Good for an overview-plus-detail figure.

```{r}
main <- vplot(mtcars) |> mark_point(x = wt, y = mpg)
mini <- vplot(mtcars) |> mark_histogram(x = mpg, bins = 8)
inset(main, mini, left = 0.6, bottom = 0.6, right = 0.98, top = 0.98)
```

### Repeat over a variable

`repeat_()` builds a small multiple by repeating one plot template across a set
of values, the composition-side counterpart to faceting when you want each
panel to be a fully independent plot rather than a shared-scale small multiple.

Faceting and composition also compose with each other: a faceted plot is still
a spec, so it can go straight into a `concat()`. Once your figure looks right,
send it to a file the same way as any plot, with `render_plot()` (see
**[Get started](quill.html)**).
