---
title: "Flows and hierarchies"
description: >
  Plot types whose positions come from a layout, not from x/y columns: sankey
  flow diagrams with vsankey (and, later, hierarchies). Axis-free, computed
  R-side, drawn through vellum primitives.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
library(vellumplot)
```

Some plot types don't map columns to `x` and `y` at all — their geometry comes
from a *layout* computed from the data's structure. `vgraph()` (see
[Spatial and networks](spatial-and-networks.html)) is one; this article covers
**flow diagrams**. Like `vgraph()`, these are whole plot types with their own
constructor, an axis-free panel, and a layout run in R before anything is drawn.

## Sankey diagrams

`vsankey()` draws a layered flow diagram from a **flow list**: one row per flow,
with a `from` node, a `to` node, and a `value` that sets the ribbon's width.

```{r}
flows <- data.frame(
  from  = c("Coal", "Gas", "Coal", "Solar", "Grid", "Grid"),
  to    = c("Grid", "Grid", "Export", "Grid", "Homes", "Industry"),
  value = c(30, 20, 10, 15, 40, 25)
)
vsankey(flows, from, to, value)
```

Nodes are the union of the `from` and `to` values; a node that appears as both a
target and a source (here, `Grid`) makes the diagram multi-stage. Columns are
placed by a longest-path layering of the flows, each node's height is
proportional to the larger of its total in- and out-flow, and ribbon widths are
proportional to `value` — so the diagram reads as a conserved flow left to right.

The flows must form a directed acyclic graph (a node cannot reach itself), and
`value`s must be positive. Turn node labels off with `label = FALSE`:

```{r}
vsankey(flows, from, to, value, label = FALSE)
```

`vsankey()` returns an ordinary [PlotSpec], so it renders with `render_plot()` /
`print()` like any other plot. Under the hood it adds a single `mark_sankey()`
layer to an axis-free, free-aspect panel; `mark_sankey()` is exported too, for
building a flow layer on a plot you have set up yourself.

## Sunburst diagrams

`vsunburst()` draws a hierarchy as concentric rings of sectors: depth maps to a
ring, and each node's angular span is its share of its parent's. Input is a
**parent list** — `id`, `parent` (`NA` for the root), and `value` (given for
leaves; an internal node's value is the sum of its children).

```{r}
h <- data.frame(
  id     = c("all", "tech", "food", "phones", "laptops", "fruit", "grain"),
  parent = c(NA, "all", "all", "tech", "tech", "food", "food"),
  value  = c(NA, NA, NA, 8, 5, 6, 4)
)
vsunburst(h, id, parent, value)
```

The root sits at the centre (it is not drawn as a wedge); the first level forms
the innermost ring. `inner_radius` opens a hole for a ring/donut sunburst:

```{r}
vsunburst(h, id, parent, value, inner_radius = 0.4)
```

Both `vsankey()` and `vsunburst()` follow the `vgraph()` pattern: the layout is
computed in R and drawn through vellum primitives, so they are ordinary specs you
can render, print, or (in future) make interactive through the same machinery as
every other plot.
