---
title: "Flows and hierarchies"
description: >
  Plot types whose positions come from a layout, not from x/y columns: sankey
  flow diagrams with vsankey (and, later, hierarchies). Axis-free, computed
  R-side, drawn through vellum primitives.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
library(vellumplot)
```

Some plot types don't map columns to `x` and `y` at all — their geometry comes
from a *layout* computed from the data's structure. `vgraph()` (see
[Spatial and networks](spatial-and-networks.html)) is one; this article covers
**flow diagrams**. Like `vgraph()`, these are whole plot types with their own
constructor, an axis-free panel, and a layout run in R before anything is drawn.

## Sankey diagrams

`vsankey()` draws a layered flow diagram from a **flow list**: one row per flow,
with a `from` node, a `to` node, and a `value` that sets the ribbon's width.

```{r}
flows <- data.frame(
  from  = c("Coal", "Gas", "Coal", "Solar", "Grid", "Grid"),
  to    = c("Grid", "Grid", "Export", "Grid", "Homes", "Industry"),
  value = c(30, 20, 10, 15, 40, 25)
)
vsankey(flows, from, to, value)
```

Nodes are the union of the `from` and `to` values; a node that appears as both a
target and a source (here, `Grid`) makes the diagram multi-stage. Columns are
placed by a longest-path layering of the flows, and nodes within each column are
ordered to minimise ribbon crossings (a deterministic barycenter sweep). Each
node's height is proportional to the larger of its total in- and out-flow, and
ribbon widths are proportional to `value` — so the diagram reads as a conserved
flow left to right.

Ribbons are coloured by their source node by default; `flow_color = "target"`
colours them by their target instead, and `flow_color = "gradient"` fades each
ribbon from its source colour to its target colour. `show_values = TRUE` appends
each node's value to its label, and `node_width` / `node_gap` tune the node
rectangles.

```{r}
vsankey(flows, from, to, value, show_values = TRUE, flow_color = "target")
```

```{r}
vsankey(flows, from, to, value, flow_color = "gradient")
```

The flows must form a directed acyclic graph (a node cannot reach itself), and
`value`s must be positive. Turn node labels off with `label = FALSE`:

```{r}
vsankey(flows, from, to, value, label = FALSE)
```

`vsankey()` returns an ordinary [PlotSpec], so it renders with `render_plot()` /
`print()` like any other plot. Under the hood it adds a single `mark_sankey()`
layer to an axis-free, free-aspect panel; `mark_sankey()` is exported too, for
building a flow layer on a plot you have set up yourself.

## Chord diagrams

`vchord()` wraps the same kind of weighted flow onto a circle: one arc (*sector*)
per node, sized by the node's total incident weight, and one ribbon per flow
joining a slice of the source sector to a slice of the target sector through the
centre. Input is a flow list (`from`, `to`, `value`) or a square flow matrix.

```{r}
flows <- data.frame(
  from  = c("Coal", "Gas", "Grid", "Grid", "Solar", "Grid"),
  to    = c("Grid", "Grid", "Homes", "Industry", "Grid", "Grid"),
  value = c(30, 20, 34, 26, 12, 6)
)
vchord(flows, from, to, value)
```

Flows are directed: each node's arc splits into an outgoing block then an
incoming block, so `a -> b` and `b -> a` are distinct ribbons and a self-flow
loops from a node's out-block to its own in-block. Sectors take a qualitative
palette; ribbons inherit their source node's colour (`link_color = "target"` to
colour by target instead). `sort = "value"` orders sectors by weight, and `gap`
sets the spacing between them:

```{r}
vchord(flows, from, to, value, sort = "value", link_color = "target")
```

## Hierarchy diagrams

`vhierarchy()` draws a tree as a space-filling diagram. All four geometries take
the same **parent list** — `id`, `parent` (`NA` for the root), and `value` (given
for leaves; an internal node's value is the sum of its children) — so you switch
between them with `type` and nothing else changes.

```{r}
h <- data.frame(
  id     = c("all", "tech", "food", "phones", "laptops", "fruit", "grain"),
  parent = c(NA, "all", "all", "tech", "tech", "food", "food"),
  value  = c(NA, NA, NA, 8, 5, 6, 4)
)
vhierarchy(h, id, parent, value) # type = "sunburst" (default)
```

The root is structural and never drawn; every top-level branch takes a distinct
hue, and its descendants inherit that hue lightened with depth, so branches stay
distinguishable and each level reads as a shade of its parent. Each node is
labelled with its `id` where the label fits (small nodes are left unlabelled, so
a dense diagram stays legible), and `show_values` appends the value.

**Sunburst** lays depth out as concentric rings, each node's angular span its
share of its parent's, starting at twelve o'clock and winding clockwise. Labels
are oriented to fit their wedge and kept upright; `root_label` writes the root
in the centre and `inner_radius` opens a hole for a ring/donut look.

```{r}
vhierarchy(h, id, parent, value, show_values = TRUE, root_label = TRUE)
```

**Treemap** packs the tree into squarified nested rectangles, and **circlepack**
into circles enclosed in circles — both read area as value:

```{r}
vhierarchy(h, id, parent, value, type = "treemap", show_values = TRUE)
vhierarchy(h, id, parent, value, type = "circlepack", show_values = TRUE)
```

**Icicle** is the rectangular cousin of the sunburst: depth becomes adjacent
bands and `flow` chooses the direction (`"down"`, `"up"`, `"right"`, `"left"`):

```{r}
vhierarchy(h, id, parent, value, type = "icicle", flow = "right")
```

### Colour

By default each top-level branch takes a hue and its descendants inherit it
lightened one step per level. The branch is an ordinary discrete fill scale, so
`scale_fill_*()` recolours the branches and `lighten` controls the fade (`0` for
a flat colour per branch):

```{r}
vhierarchy(h, id, parent, value, type = "treemap", lighten = 0.3) |>
  scale_fill_manual(values = c(tech = "#1b7837", food = "#762a83"))
```

Map `fill` to a node column to colour every node by that variable instead, with
the matching `scale_fill_*()`. The depth fade is dropped, since the colour now
carries its own meaning:

```{r}
h$owner <- c(NA, NA, NA, "Sam", "Sam", "Lee", "Lee")
vhierarchy(h, id, parent, value, fill = owner, type = "circlepack")
```

`vsankey()`, `vchord()`, and `vhierarchy()` all follow the `vgraph()` pattern: the layout is
computed in R and drawn through vellum primitives, so they are ordinary specs you
can render, print, or (in future) make interactive through the same machinery as
every other plot.
