---
title: "Declarative interactivity"
description: >
  Declare interaction as part of the plot: named selections bound to gestures,
  conditional encodings that style by membership, filters that show only the
  selection, and cross-view cross-filtering. Inert on a static render; enacted by
  a host such as vellumwidget.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

Most of vellumplot describes a *static* picture. But interaction can be part of
the plot too — declared in the spec, travelling with it, and enacted by any
capable host (today, [vellumwidget](https://r-vellum.github.io/vellumwidget/)).
This is different from configuring a widget after the fact: the interaction is a
first-class, serialisable piece of the plot.

Everything on this page is **inert on a static render** — a plot with
interactions compiles and draws exactly like one without (the figures below are
the static renders). The interaction comes alive only when a host such as
`vellumwidget::as_widget()` enacts it.

## Selections

A **selection** is a named set of data elements defined by a user gesture. On its
own it does nothing; you refer to it elsewhere. Declare one with `select_point()`
(a click or hover, or a legend value) or `select_interval()` (a brush or lasso,
optionally locked to an axis):

```{r}
#| label: selection
#| eval: false
p |> select_point("hi", on = "hover") # hover a mark
p |> select_interval("brush", on = "x") # drag an x-range
```

`on` sets the gesture; `empty = TRUE` (the default) means an *empty* selection
contains everything — so an untouched plot shows its full self and interacting
*narrows* it.

## Conditional encoding: `condition()`

Use `condition()` as the value of an aesthetic to make it depend on selection
membership: members get the first value, non-members the second.

```{r}
#| label: condition
df <- data.frame(x = mtcars$wt, y = mtcars$mpg, cyl = factor(mtcars$cyl))
vplot(df) |>
  mark_point(x = x, y = y, color = condition("hi", cyl, "grey80")) |>
  select_point("hi", on = "hover")
```

`condition()` is transparent to the grammar: the `if_true` branch trains the
colour scale and draws the legend exactly as `color = cyl` would, so the static
render is the full, meaningful plot. With `empty = TRUE`, nothing is selected
initially, so every point shows its `if_true` colour; once a host activates the
selection, non-members switch to `if_false` (here `"grey80"`) — the spotlight.

Omit `if_false` to fall back to the theme's dim appearance:

```{r}
#| label: condition-dim
#| eval: false
mark_point(x = x, y = y, color = condition("hi", cyl)) # non-members just dim
```

## Filtering: `filter_by()`

`filter_by()` shows only the selection's members, hiding the rest:

```{r}
#| label: filter
#| eval: false
vplot(df) |>
  mark_point(x = x, y = y) |>
  select_interval("brush", on = "xy") |>
  filter_by("brush")
```

### Cross-filtering across views

The real power is pointing a *second* view at a selection defined on a *first* —
brush one panel, a linked panel narrows to those rows while the source stays
full. Define the selection once, `add_selection()` it on the source, and
`filter_by()` it on the target:

```{r}
#| label: crossfilter
sel <- select_interval("brush", on = "xy")
hconcat(
  vplot(df) |> mark_point(x = x, y = y) |> add_selection(sel),
  vplot(df) |> mark_point(x = y, y = x) |> filter_by(sel)
)
```

The two views share row identity, so a gesture in one maps to the matching rows
in the other. This cross-view coordination is exactly what a widget *flag* cannot
express — it lives in the plot.

## Overview + detail: `bind_scale()`

`bind_scale()` binds a panel's view to an interval selection on another (an
overview), so brushing the overview pans/zooms the detail. It is declared in the
spec today; host enactment is in progress.

## Which marks are addressable

Give a mark an identity with `data_id=` (and, optionally, a `tooltip=`) and it
becomes hoverable, selectable and cross-filterable. This is no longer limited to
the batched marks (points, bars, tiles, segments): a line or step **series** is
one addressable object, a filled `mark_area()` / `mark_ribbon()` band is one
polygon, `mark_text()` labels are addressable per datum, `mark_label()` keys by
its rounded background box, and an sf polygon / choropleth keys per feature.

```{r}
#| label: keyed-line
#| eval: false
# hover the whole trend line as one object
vplot(pressure) |>
  mark_line(x = temperature, y = pressure, data_id = "vapor-pressure") |>
  select_point("series", on = "hover")
```

A mark you leave without a key is simply not addressable — and, for lines,
polygons and text, it contributes no row to the interaction model at all. That
is deliberate: a plot's gridlines and axis labels are unkeyed text and lines, and
they must not drown the marks that carry meaning. So *absence of a key is a
choice you are making*, not a limitation.

## What a host does with it

`vellumplot` compiles once and carries the interaction declarations plus the
per-element metadata a host needs (`interaction_model()` returns the declaration
block). The host reads them and wires the gestures it already performs on the
frozen scene — highlight, hide, pan/zoom — with no recompilation. Turn any of the
plots above into a live widget with:

```{r}
#| label: as-widget
#| eval: false
library(vellumwidget)
vplot(df) |>
  mark_point(x = x, y = y, color = condition("hi", cyl, "grey80")) |>
  select_point("hi", on = "hover") |>
  as_widget()
```

The interaction is display-tier by design: it reacts on the frozen scene
(highlight, filter-by-hiding, pan/zoom) and never recomputes the grammar. See
`select_point()`, `select_interval()`, `condition()`, `filter_by()`,
`add_selection()`, and `bind_scale()` in the reference.
</content>
