---
title: "Marks, layer by layer"
description: >
  A tour of the mark families: points, lines and bars, areas and intervals,
  tiles and bins, text, images, and pie. How encodings map, and how scales
  train across every layer on the panel.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A vellumplot plot is a stack of *marks*. Each `mark_*()` appends one drawing layer
to the spec, and every mark reads the same grammar: bare column names (or
expressions) captured with tidy evaluation become *encodings*, while scalar
values become constant aesthetics. `color = hp` maps the `hp` column through a
scale; `color = "red"` paints every element red.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3)
```

You can stack as many marks as you like on a panel, and scales train across
all of them at once.

## Points, lines, and bars

`mark_point()` draws markers; `mark_line()` and `mark_step()` connect points in
`x` order; `mark_rule()` draws reference lines. `mark_point()` also takes
`shape` (one of `"circle"`, `"square"`, `"triangle"`, `"diamond"`, `"plus"`,
`"cross"`) and a `position = "jitter"` adjustment.

```{r}
vplot(pressure) |>
  mark_line(x = temperature, y = pressure) |>
  mark_point(x = temperature, y = pressure)
```

`shape` also accepts an **SVG icon** — a path `d` string (what icon sets ship)
or a `.svg` file — drawn as a crisp vector marker. Use a literal string for a
constant icon, or map `shape` and give [scale_shape()] one icon per level (the
legend shows the icons); `size` scales them.

```{r}
star <- "M12 2l3 7h7l-5.5 4.5 2 7-6.5-4.5-6.5 4.5 2-7L2 9h7z"
heart <- "M12 21s-7-4.35-9.5-8.5C1 9 2.5 5 6 5c2 0 3.5 1.5 4 2.5C10.5 6.5 12 5 14 5c3.5 0 5 4 3.5 7.5C19 16.65 12 21 12 21z"
d <- data.frame(x = 1:6, y = c(2, 4, 3, 5, 4, 6), g = rep(c("a", "b"), 3))
vplot(d) |>
  mark_point(x = x, y = y, shape = g, color = g, size = 1.6) |>
  scale_shape(values = c(a = heart, b = star))
```

`mark_bar()` draws bars from a zero baseline. Give it an explicit `y` for
heights, or omit `y` and it counts rows per category (the count stat). When
`fill` is mapped, groups stack by default; switch to side-by-side with
`position = "dodge"` or normalise to 1 with `position = "fill"`.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(gear), position = "dodge")
```

For finer control, the `position` argument also takes a parameterised
`position_*()` object. `position_dodge2()` fills each category's band by the
groups actually *present* (so a ragged grouping stays centred); `position_nudge()`
offsets by a constant; `position_jitter(width=, seed=)` and
`position_jitterdodge()` control scatter for overplotted categorical points.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(gear), position = position_dodge2(padding = 0.15))
```

## Areas and intervals

`mark_area()` fills between a `y` line and zero, `mark_ribbon()` fills between
`ymin` and `ymax`, and the interval marks draw ranges: `mark_errorbar()` (with
caps), `mark_linerange()` (without), and `mark_segment()` from `(x, y)` to
`(xend, yend)`.

```{r}
vplot(pressure) |>
  mark_area(x = temperature, y = pressure, fill = "steelblue", alpha = 0.4) |>
  mark_line(x = temperature, y = pressure)
```

`mark_boxplot()` summarises the raw `y` values per `x` category into a
box-and-whisker (box from Q1 to Q3, median line, whiskers at 1.5 times the
IQR, outliers as points).

```{r}
vplot(mtcars) |>
  mark_boxplot(x = factor(cyl), y = mpg)
```

Like points and bars, these summary marks are *addressable*: an error bar or line
range keyed with `data_id`/`tooltip` carries that identity on every segment it
draws, and a boxplot keys each box by its category — so they hover, tooltip, and
select as units once rendered as an interactive widget.

`mark_pointrange()` and `mark_crossbar()` are the *identity* summary marks: you
supply `y`, `ymin`, and `ymax` (e.g. a model's estimate and interval) and they
draw a point-with-range or a box-with-centre-line — no aggregation.

## Reference lines and function curves

`mark_abline()` draws a sloped reference line `y = slope * x + intercept`, the
diagonal companion to `mark_rule()`'s horizontal and vertical lines. Both
`slope` and `intercept` may be vectors to draw a family of lines.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_abline(slope = -5, intercept = 37, color = "firebrick")
```

`mark_function()` draws a curve `y = fun(x)` sampled across the panel's x range —
handy for eyeballing a theoretical distribution against the data. Overlay a
normal density on a density histogram:

```{r}
set.seed(1)
vplot(data.frame(z = rnorm(500))) |>
  mark_histogram(x = z, y = after_stat(density)) |>
  mark_function(fun = dnorm, color = "firebrick", linewidth = 1.5)
```

Both reuse the panel's existing scales, so add them on top of a data layer.
(`mark_function()` holds a live function, so it does not round-trip through
`as_spec()`.)

When many observations land on the same `(x, y)` — common with rounded or
discrete data — `mark_count()` collapses them to one bubble sized by the overlap
count, an honest alternative to invisible overplotting:

```{r}
vplot(data.frame(cyl = mtcars$cyl, gear = mtcars$gear)) |>
  mark_count(x = cyl, y = gear)
```

## Tiles and bins

`mark_tile()` draws a rectangle at each `(x, y)` coloured by `fill`;
`mark_raster()` is the same thing drawn as a single raster image (a fast path
that needs a complete regular grid). For continuous data, `mark_bin2d()` and
`mark_hex()` bin `x` and `y` into a grid and colour each cell by count.

```{r}
grid <- expand.grid(x = 1:8, y = 1:8)
grid$z <- with(grid, sin(x / 2) + cos(y / 2))
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_continuous(palette = "Batlow")
```

`mark_contour()` draws iso-density contour lines of a 2-D point cloud (and
`mark_contour_filled()` fills the bands), coloured by level. See the
[statistical marks](statistical-marks.html) article for the details.

```{r}
vplot(faithful) |>
  mark_point(x = eruptions, y = waiting, color = "grey70") |>
  mark_contour(x = eruptions, y = waiting)
```

## Text

`mark_text()` draws the `label` aesthetic as text at each `(x, y)`;
`mark_label()` adds a filled background behind each label so it stays legible
over busy marks. `size` is in points, and `angle` can be mapped or constant.

```{r}
top <- mtcars[mtcars$mpg > 30, ]
vplot(top) |>
  mark_point(x = wt, y = mpg) |>
  mark_text(x = wt, y = mpg, label = rownames(top), vjust = "bottom", size = 9)
```

On a crowded scatter, labels collide. `repel = TRUE` moves them apart and draws
a thin leader back to each point. Placement is solved by the engine in device
pixels and applied as an absolute offset, so it does not drift with the data
scale and is deterministic (no seed needed). Because the offset is coordinate
agnostic, repel also works on **faceted and polar** plots — each panel is solved
and its labels kept inside it.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_text(
    x = wt, y = mpg, label = rownames(mtcars),
    repel = TRUE, size = 4
  )
```

`mark_text_path()` sets a label *along* a curve instead of at a point -- one
label per group, its glyphs following the group's path and rotated to the local
tangent. It is the direct way to label a line without a legend, or to caption a
contour or arc. Glyphs follow the tangent, so a path walked right-to-left reads
upside-down -- traverse it in the reading direction (here `sin` over a rising
half-period).

```{r}
t <- seq(0, pi, length.out = 60)
curve <- data.frame(x = t, y = sin(t), lab = "y = sin(x)")
vplot(curve) |>
  mark_text_path(x = x, y = y, label = lab, size = 9, vjust = "bottom", offset = 2)
```

`mark_series_label()` is the other legend-free way to name lines: it puts each
series' name at its **end** (the point with the largest `x`), coloured to match
and repelled apart. Map the same `x`/`y`/`color` as the lines — the label text
and colour follow the series. Give the panel a little x-room and drop the
now-redundant colour legend.

```{r}
set.seed(1)
econ <- data.frame(
  year = rep(2000:2015, 3),
  value = c(
    cumsum(rnorm(16, 2)), cumsum(rnorm(16, 1)), cumsum(rnorm(16, 3))
  ),
  series = rep(c("north", "south", "east"), each = 16)
)
vplot(econ) |>
  mark_line(x = year, y = value, color = series) |>
  mark_series_label(x = year, y = value, color = series) |>
  xlim(2000, 2018) |>
  guides(color = "none")
```

`mark_outlier_label()` labels only the points that *stand out* — it keeps the
rows whose `y` is an outlier (Tukey's IQR rule, or `method = "sd"`) and labels
just those, so a busy scatter names its extremes without a wall of text. Map a
`label` to name each outlier; with a `color`/`fill` mapped, outliers are found
within each group.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_outlier_label(x = wt, y = mpg, label = rownames(mtcars))
```

## Images

`mark_image()` puts a bitmap at each `(x, y)` in place of a marker: a flag or a
company logo. `src` is a column of file paths (one image per datum) or a single
path reused at every point. `size` sets the height in millimetres, and the
width follows each image's own aspect ratio, so nothing stretches.

```{r}
#| eval: !expr requireNamespace("vellumplot", quietly = TRUE) && requireNamespace("magick", quietly = TRUE)
badge <- function(text, fill) {
  path <- tempfile(fileext = ".png")
  magick::image_blank(120, 120, color = fill) |>
    magick::image_annotate(text, size = 64, gravity = "center", color = "white") |>
    magick::image_write(path)
  path
}
d <- data.frame(
  x = 1:3,
  y = c(2, 3, 1),
  logo = c(badge("A", "tomato"), badge("B", "steelblue"), badge("C", "seagreen"))
)
vplot(d) |>
  mark_image(x = x, y = y, src = logo, size = 14)
```

Reading images needs the magick package (a suggested dependency), which decodes
PNG, JPEG, SVG, and more. Because `size` is in millimetres rather than data
units, images keep their physical size as the panel resizes.

## Pie and donut

`mark_pie()` and `mark_donut()` are the part-of-whole shortcuts. Each `value`
becomes a wedge; `fill` colours the slices. Under the hood they are a stacked
bar projected through `coord_polar()`, which they set for you.

```{r}
parts <- data.frame(part = c("a", "b", "c", "d"), n = c(3, 5, 2, 4))
vplot(parts) |>
  mark_donut(value = n, fill = part, inner_radius = 0.6)
```

For polar plots generally, `coord_radial()` extends `coord_polar()` with a
central hole (`inner_radius`) and a partial `start`–`end` arc — e.g. a
semicircular coxcomb:

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(cyl)) |>
  coord_radial(theta = "x", start = -pi / 2, end = pi / 2, inner_radius = 0.2)
```

## Layering is the point

Because scales train across every layer, mixing marks on one panel works.
Here a point cloud and a fitted line share the same trained `x` and `y` axes.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

From here:

- **[Scales and guides](scales-and-guides.html)** for the mapping from data to
  colour, size, and axes.
- **[Statistical marks](statistical-marks.html)** for histograms, densities,
  and smooths in depth.
- **[Spatial and networks](spatial-and-networks.html)** for `mark_sf()` maps
  and `vgraph()` node-link diagrams.
