---
title: "Marks, layer by layer"
description: >
  A tour of the mark families: points, lines and bars, areas and intervals,
  tiles and bins, text, and pie. How encodings map, and how scales train
  across every layer on the panel.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A vellumplot plot is a stack of *marks*. Each `mark_*()` appends one drawing layer
to the spec, and every mark reads the same grammar: bare column names (or
expressions) captured with tidy evaluation become *encodings*, while scalar
values become constant aesthetics. `color = hp` maps the `hp` column through a
scale; `color = "red"` paints every element red.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3)
```

The rest of this article walks the mark families. The one idea to hold onto:
you can stack as many marks as you like on a panel, and scales train across
all of them at once.

## Points, lines, and bars

`mark_point()` draws markers; `mark_line()` and `mark_step()` connect points in
`x` order; `mark_rule()` draws reference lines. `mark_point()` also takes
`shape` (one of `"circle"`, `"square"`, `"triangle"`, `"diamond"`, `"plus"`,
`"cross"`) and a `position = "jitter"` adjustment.

```{r}
vplot(pressure) |>
  mark_line(x = temperature, y = pressure) |>
  mark_point(x = temperature, y = pressure)
```

`mark_bar()` draws bars from a zero baseline. Give it an explicit `y` for
heights, or omit `y` and it counts rows per category (the count stat). When
`fill` is mapped, groups stack by default; switch to side-by-side with
`position = "dodge"` or normalise to 1 with `position = "fill"`.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(gear), position = "dodge")
```

## Areas and intervals

`mark_area()` fills between a `y` line and zero, `mark_ribbon()` fills between
`ymin` and `ymax`, and the interval marks draw ranges: `mark_errorbar()` (with
caps), `mark_linerange()` (without), and `mark_segment()` from `(x, y)` to
`(xend, yend)`.

```{r}
vplot(pressure) |>
  mark_area(x = temperature, y = pressure, fill = "steelblue", alpha = 0.4) |>
  mark_line(x = temperature, y = pressure)
```

`mark_boxplot()` summarises the raw `y` values per `x` category into a
box-and-whisker (box from Q1 to Q3, median line, whiskers at 1.5 times the
IQR, outliers as points).

```{r}
vplot(mtcars) |>
  mark_boxplot(x = factor(cyl), y = mpg)
```

## Tiles and bins

`mark_tile()` draws a rectangle at each `(x, y)` coloured by `fill`;
`mark_raster()` is the same thing drawn as a single raster image (a fast path
that needs a complete regular grid). For continuous data, `mark_bin2d()` and
`mark_hex()` bin `x` and `y` into a grid and colour each cell by count.

```{r}
grid <- expand.grid(x = 1:8, y = 1:8)
grid$z <- with(grid, sin(x / 2) + cos(y / 2))
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_continuous(palette = "Batlow")
```

`mark_contour()` draws iso-density contour lines of a 2-D point cloud (and
`mark_contour_filled()` fills the bands), coloured by level — see the
[statistical marks](statistical-marks.html) article for the details.

```{r}
vplot(faithful) |>
  mark_point(x = eruptions, y = waiting, color = "grey70") |>
  mark_contour(x = eruptions, y = waiting)
```

## Text

`mark_text()` draws the `label` aesthetic as text at each `(x, y)`;
`mark_label()` adds a filled background behind each label so it stays legible
over busy marks. `size` is in points, and `angle` can be mapped or constant.

```{r}
top <- mtcars[mtcars$mpg > 30, ]
vplot(top) |>
  mark_point(x = wt, y = mpg) |>
  mark_text(x = wt, y = mpg, label = rownames(top), vjust = "bottom", size = 9)
```

## Pie and donut

`mark_pie()` and `mark_donut()` are the part-of-whole shortcuts. Each `value`
becomes a wedge; `fill` colours the slices. Under the hood they are a stacked
bar projected through `coord_polar()`, which they set for you.

```{r}
parts <- data.frame(part = c("a", "b", "c", "d"), n = c(3, 5, 2, 4))
vplot(parts) |>
  mark_donut(value = n, fill = part, hole = 0.6)
```

## Layering is the point

Because scales train across every layer, mixing marks on one panel just works.
Here a point cloud and a fitted line share the same trained `x` and `y` axes.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

From here:

- **[Scales and guides](scales-and-guides.html)** for the mapping from data to
  colour, size, and axes.
- **[Statistical marks](statistical-marks.html)** for histograms, densities,
  and smooths in depth.
- **[Spatial and networks](spatial-and-networks.html)** for `mark_sf()` maps
  and `vgraph()` node-link diagrams.
