---
title: "Marks, layer by layer"
description: >
  A tour of the mark families: points, lines and bars, areas and intervals,
  tiles and bins, text, images, and pie. How encodings map, and how scales
  train across every layer on the panel.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A vellumplot plot is a stack of *marks*. Each `mark_*()` appends one drawing layer
to the spec, and every mark reads the same grammar: bare column names (or
expressions) captured with tidy evaluation become *encodings*, while scalar
values become constant aesthetics. `color = hp` maps the `hp` column through a
scale; `color = "red"` paints every element red.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3)
```

You can stack as many marks as you like on a panel, and scales train across
all of them at once.

## Points, lines, and bars

`mark_point()` draws markers; `mark_line()` and `mark_step()` connect points in
`x` order; `mark_rule()` draws reference lines. `mark_point()` also takes
`shape` (one of `"circle"`, `"square"`, `"triangle"`, `"diamond"`, `"plus"`,
`"cross"`) and a `position = "jitter"` adjustment.

```{r}
vplot(pressure) |>
  mark_line(x = temperature, y = pressure) |>
  mark_point(x = temperature, y = pressure)
```

`mark_bar()` draws bars from a zero baseline. Give it an explicit `y` for
heights, or omit `y` and it counts rows per category (the count stat). When
`fill` is mapped, groups stack by default; switch to side-by-side with
`position = "dodge"` or normalise to 1 with `position = "fill"`.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(gear), position = "dodge")
```

For finer control, the `position` argument also takes a parameterised
`position_*()` object. `position_dodge2()` fills each category's band by the
groups actually *present* (so a ragged grouping stays centred); `position_nudge()`
offsets by a constant; `position_jitter(width=, seed=)` and
`position_jitterdodge()` control scatter for overplotted categorical points.

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(gear), position = position_dodge2(padding = 0.15))
```

## Areas and intervals

`mark_area()` fills between a `y` line and zero, `mark_ribbon()` fills between
`ymin` and `ymax`, and the interval marks draw ranges: `mark_errorbar()` (with
caps), `mark_linerange()` (without), and `mark_segment()` from `(x, y)` to
`(xend, yend)`.

```{r}
vplot(pressure) |>
  mark_area(x = temperature, y = pressure, fill = "steelblue", alpha = 0.4) |>
  mark_line(x = temperature, y = pressure)
```

`mark_boxplot()` summarises the raw `y` values per `x` category into a
box-and-whisker (box from Q1 to Q3, median line, whiskers at 1.5 times the
IQR, outliers as points).

```{r}
vplot(mtcars) |>
  mark_boxplot(x = factor(cyl), y = mpg)
```

Like points and bars, these summary marks are *addressable*: an error bar or line
range keyed with `data_id`/`tooltip` carries that identity on every segment it
draws, and a boxplot keys each box by its category — so they hover, tooltip, and
select as units once rendered as an interactive widget.

## Tiles and bins

`mark_tile()` draws a rectangle at each `(x, y)` coloured by `fill`;
`mark_raster()` is the same thing drawn as a single raster image (a fast path
that needs a complete regular grid). For continuous data, `mark_bin2d()` and
`mark_hex()` bin `x` and `y` into a grid and colour each cell by count.

```{r}
grid <- expand.grid(x = 1:8, y = 1:8)
grid$z <- with(grid, sin(x / 2) + cos(y / 2))
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_continuous(palette = "Batlow")
```

`mark_contour()` draws iso-density contour lines of a 2-D point cloud (and
`mark_contour_filled()` fills the bands), coloured by level. See the
[statistical marks](statistical-marks.html) article for the details.

```{r}
vplot(faithful) |>
  mark_point(x = eruptions, y = waiting, color = "grey70") |>
  mark_contour(x = eruptions, y = waiting)
```

## Text

`mark_text()` draws the `label` aesthetic as text at each `(x, y)`;
`mark_label()` adds a filled background behind each label so it stays legible
over busy marks. `size` is in points, and `angle` can be mapped or constant.

```{r}
top <- mtcars[mtcars$mpg > 30, ]
vplot(top) |>
  mark_point(x = wt, y = mpg) |>
  mark_text(x = wt, y = mpg, label = rownames(top), vjust = "bottom", size = 9)
```

On a crowded scatter, labels collide. `repel = TRUE` moves them apart with a
force-directed layout and draws a thin leader back to each point. The layout is
resolved against the true rendered panel size (so it doesn't drift with the data
scale) and is reproducible via `seed`.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_text(
    x = wt, y = mpg, label = rownames(mtcars),
    repel = TRUE, size = 4, seed = 1
  )
```

`mark_text_path()` sets a label *along* a curve instead of at a point -- one
label per group, its glyphs following the group's path and rotated to the local
tangent. It is the direct way to label a line without a legend, or to caption a
contour or arc. Glyphs follow the tangent, so a path walked right-to-left reads
upside-down -- traverse it in the reading direction (here `sin` over a rising
half-period).

```{r}
t <- seq(0, pi, length.out = 60)
curve <- data.frame(x = t, y = sin(t), lab = "y = sin(x)")
vplot(curve) |>
  mark_text_path(x = x, y = y, label = lab, size = 9, vjust = "bottom", offset = 2)
```

## Images

`mark_image()` puts a bitmap at each `(x, y)` in place of a marker: a flag or a
company logo. `src` is a column of file paths (one image per datum) or a single
path reused at every point. `size` sets the height in millimetres, and the
width follows each image's own aspect ratio, so nothing stretches.

```{r}
#| eval: !expr requireNamespace("vellumplot", quietly = TRUE) && requireNamespace("magick", quietly = TRUE)
badge <- function(text, fill) {
  path <- tempfile(fileext = ".png")
  magick::image_blank(120, 120, color = fill) |>
    magick::image_annotate(text, size = 64, gravity = "center", color = "white") |>
    magick::image_write(path)
  path
}
d <- data.frame(
  x = 1:3,
  y = c(2, 3, 1),
  logo = c(badge("A", "tomato"), badge("B", "steelblue"), badge("C", "seagreen"))
)
vplot(d) |>
  mark_image(x = x, y = y, src = logo, size = 14)
```

Reading images needs the magick package (a suggested dependency), which decodes
PNG, JPEG, SVG, and more. Because `size` is in millimetres rather than data
units, images keep their physical size as the panel resizes.

## Pie and donut

`mark_pie()` and `mark_donut()` are the part-of-whole shortcuts. Each `value`
becomes a wedge; `fill` colours the slices. Under the hood they are a stacked
bar projected through `coord_polar()`, which they set for you.

```{r}
parts <- data.frame(part = c("a", "b", "c", "d"), n = c(3, 5, 2, 4))
vplot(parts) |>
  mark_donut(value = n, fill = part, inner_radius = 0.6)
```

For polar plots generally, `coord_radial()` extends `coord_polar()` with a
central hole (`inner_radius`) and a partial `start`–`end` arc — e.g. a
semicircular coxcomb:

```{r}
vplot(mtcars) |>
  mark_bar(x = factor(cyl), fill = factor(cyl)) |>
  coord_radial(theta = "x", start = -pi / 2, end = pi / 2, inner_radius = 0.2)
```

## Layering is the point

Because scales train across every layer, mixing marks on one panel works.
Here a point cloud and a fitted line share the same trained `x` and `y` axes.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

From here:

- **[Scales and guides](scales-and-guides.html)** for the mapping from data to
  colour, size, and axes.
- **[Statistical marks](statistical-marks.html)** for histograms, densities,
  and smooths in depth.
- **[Spatial and networks](spatial-and-networks.html)** for `mark_sf()` maps
  and `vgraph()` node-link diagrams.
