---
title: "Scales and guides"
description: >
  How data values become colour, size, shape, opacity, line type, and axis
  positions. Continuous, discrete, binned, manual, gradient, date/time, and
  identity scales, limit shortcuts, trained domains, and controlling the legends
  they produce.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
library(vellumplot)
```

A *scale* is the function that turns a data value into something you can see: a
colour, a marker size, a shape, or a position along an axis. In vellumplot every
mapped encoding gets a scale automatically, trained from the data, so most
plots need no explicit scale at all. You reach for a `scale_*()` call only when
you want to override the default: change the palette, set a transform, fix the
limits, or rename the guide.

The key idea is *training*. When you map `color = hp`, vellumplot scans the `hp`
values across every layer, works out the domain, and builds a continuous colour
scale with a legend. Add a `scale_*()` and you are declaring an override on top
of that trained default, not replacing the whole machine.

## Colour, continuous and discrete

Continuous data get a smooth colour ramp; discrete data get a categorical
palette. The `palette` argument takes a vector of colours or a single palette
name passed to `grDevices::hcl.colors()` (for example `"Batlow"`, `"Blues"`,
`"Set 2"`).

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3) |>
  scale_color_continuous(palette = "Viridis", name = "Horsepower")
```

For a two-colour ramp, `scale_color_gradient()` takes the endpoints directly.
For categories, `scale_color_manual()` maps levels to colours; naming the
values pins each level to a specific colour.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl), size = 3) |>
  scale_color_manual(values = c("4" = "#1b9e77", "6" = "#d95f02", "8" = "#7570b3"))
```

For signed or anomaly data — where a value like zero is meaningfully neutral —
`scale_color_gradient2()` (and `scale_fill_gradient2()`) draw a diverging ramp
around a `midpoint`. It rescales *about* the midpoint, so the neutral `mid`
colour always lands on that value even when the data range is lopsided:

```{r}
anom <- data.frame(x = 1:20, y = 1:20, resid = seq(-3, 6, length.out = 20))
vplot(anom) |>
  mark_point(x = x, y = y, color = resid, size = 3) |>
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#b2182b", midpoint = 0)
```

Ramps built from a plain colour vector — a `scale_color_gradient(low, high)`, or a
`palette` given as colours — are interpolated in the perceptually-uniform **Oklab**
space, not sRGB, so the ramp and its colourbar read evenly instead of dipping
through a muddy, over-dark middle (a blue→yellow ramp no longer passes through
grey). Designed perceptual palettes such as the default `Batlow` or any
`hcl.colors()` name are already uniform and unchanged. To blend in sRGB (or CIE
Lab) instead, set `options(vellumplot.color.interpolation = "srgb")` (or `"lab"`).
A gradient *fill* opts in per gradient with
`linear_gradient(..., interpolation = "oklab")`.

## Binned colour

`scale_fill_binned()` and `scale_color_binned()` cut a continuous aesthetic
into classes and give it a discrete legend, which is what you want for a
choropleth or a heatmap you read by band rather than by exact value. Choose the
classification `style` (`"quantile"`, `"equal"`, `"pretty"`, or any
`classInt::classIntervals()` style) and the number of classes `n`.

```{r}
grid <- expand.grid(x = 1:10, y = 1:10)
grid$z <- with(grid, x * y)
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_binned(style = "quantile", n = 5, palette = "Mako")
```

## Size, shape, and edge width

Non-colour aesthetics have scales too. `scale_size()` maps values linearly to a
marker-size `range` (in mm); `scale_shape()` cycles a set of shapes over the
levels of a discrete aesthetic; `scale_edge_width()` does for network edges
what `scale_size()` does for points.

The shape palette is `"circle"`, `"square"`, `"triangle"`, `"diamond"`, `"plus"`,
`"cross"`, `"triangle_down"`, and `"star"` — eight in all, so a mapped `shape`
covers up to eight levels automatically. Pass a subset (or a reordering) with
`scale_shape(values = ...)`. Filled shapes take the mark's `fill`/`color`, so an
open marker is `fill = NA` with a `color`.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = hp, color = factor(cyl)) |>
  scale_size(range = c(2, 9), name = "hp")
```

`scale_size()` maps the value to the marker's *radius*. For bubble charts prefer
`scale_size_area()`, which maps to the *area* (value 0 → size 0) so the ink is
proportional to the value — the honest encoding:

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = hp) |>
  scale_size_area(max_size = 10)
```

## Opacity and line type

`alpha` and `linetype` are mapped aesthetics too. `scale_alpha()` maps a
continuous variable to opacity (its `range` defaults to `c(0.1, 1)`), a good way
to let density show through an overplotted cloud.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, alpha = hp, size = 3) |>
  scale_alpha(range = c(0.2, 1))
```

`scale_linetype()` maps a discrete variable to line types, cycling `"solid"`,
`"dashed"`, `"dotted"`, and so on. It applies to line-like marks
(`mark_line()`, `mark_step()`).

```{r}
series <- data.frame(
  x = rep(1:20, 3),
  y = c(cumsum(rnorm(20)), cumsum(rnorm(20)), cumsum(rnorm(20))),
  series = rep(c("a", "b", "c"), each = 20)
)
vplot(series) |>
  mark_line(x = x, y = y, linetype = series)
```

## Identity scales

Sometimes a column *already* holds the exact aesthetic values you want: actual
colour names, sizes in millimetres, shape names. An identity scale uses them
verbatim and draws no legend. There is a variant for each aesthetic:
`scale_color_identity()`, `scale_fill_identity()`, `scale_size_identity()`,
`scale_shape_identity()`, `scale_alpha_identity()`, `scale_linetype_identity()`.

```{r}
df <- data.frame(
  x = 1:5, y = 1:5,
  col = c("firebrick", "goldenrod", "forestgreen", "steelblue", "purple")
)
vplot(df) |>
  mark_point(x = x, y = y, color = col, size = 6) |>
  scale_color_identity()
```

## Position scales

`scale_x_continuous()` and `scale_y_continuous()` control the axes. By default
they train from the data with a small expansion; override them to set `limits`,
apply a `trans` (`"log10"`, `"sqrt"`, `"symlog"`, `"reverse"`), or supply
explicit `breaks` and `labels`. Categorical axes use `scale_x_discrete()` and
`scale_y_discrete()`, whose `limits` set the order or subset of levels.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  scale_y_continuous(limits = c(10, 35), breaks = seq(10, 35, 5)) |>
  scale_x_continuous(trans = "log10", name = "Weight (log scale)")
```

Where `"log10"` cannot go — data that crosses zero or spans positive and negative
values over several orders of magnitude — `"symlog"` gives a symmetric-log axis:
linear through zero, logarithmic in each tail, with ticks at zero and signed
powers of ten.

```{r}
signed <- data.frame(v = c(-1000, -100, -10, 0, 10, 100, 1000), y = 1:7)
vplot(signed) |>
  mark_point(x = v, y = y) |>
  scale_x_continuous(trans = "symlog")
```

To set limits without spelling out a whole scale, use the shortcuts `xlim()`,
`ylim()`, and `lims()` (the last takes one named argument per aesthetic):

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  lims(x = c(0, 6), y = c(0, 40))
```

### Scale transform vs display transform

A `scale_*(trans=)` (above) transforms the *data*: it picks its breaks in the
transformed space, so a `"log10"` axis is labelled `1, 10, 100`. `coord_trans()`
instead warps only the *display*, after the scale has trained — the breaks stay at
their original data values, so the axis keeps those labels but they sit at warped
positions (gridlines bunch up, and straight lines curve). Use it to show data on a
log display without relabelling the axis in powers of ten:

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_line(x = wt, y = mpg) |>
  coord_trans(y = "log10")
```

Each of `x` / `y` takes a transform name (`"log10"`, `"sqrt"`, `"identity"`) or a
`scales::transform_*()` object. It applies to the common marks (points, lines,
areas, bars, tiles, smooths, text); interval/segment, boxplot, and raster marks
are not warped yet.

### Secondary axes

A secondary axis adds a second set of ticks and labels on the opposite edge (top
for `x`, right for `y`), computed as a 1:1 monotonic transform of the primary
axis. Pass `sec_axis()` to the `sec.axis` argument of a continuous position
scale. The classic use is a unit conversion — plot in Celsius, label the top in
Fahrenheit:

```{r}
vplot(data.frame(celsius = 0:100, y = (0:100)^2)) |>
  mark_line(x = celsius, y = y) |>
  scale_x_continuous(name = "°C", sec.axis = sec_axis(~ . * 1.8 + 32, name = "°F"))
```

`dup_axis()` is the identity special case — it simply duplicates the axis on the
far edge, handy for reading a wide plot from either side:

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  scale_y_continuous(sec.axis = dup_axis())
```

The transform can be a formula (`~ . * 2`), a function, or a
`scales::transform_*()` object, and must be monotonic. This is a labelling
convenience, *not* an independent second axis with its own data. It is currently
supported on continuous position scales under the default Cartesian coordinate
system, with shared scales across facets; combining it with `coord_flip()` /
`coord_polar()` / `coord_trans()`, with free facet scales, or with
`add_marginal()` raises an error, and in a plot composition the secondary axis is
not drawn.

## Date and time axes

A `Date` or `POSIXct` column gets a date axis automatically. To control the
break interval or the label format, declare `scale_x_date()` (or
`scale_x_datetime()` / `scale_x_time()`): `date_breaks` takes an interval string
like `"6 months"`, and `date_labels` a `strftime()` format.

```{r}
econ <- data.frame(
  day = as.Date("2020-01-01") + 0:729,
  value = cumsum(rnorm(730))
)
vplot(econ) |>
  mark_line(x = day, y = value) |>
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y")
```

## Guides come for free

Every scale that needs a legend produces one, and vellumplot stacks multiple
legends automatically. Map two aesthetics and you get two guides without asking.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = disp) |>
  scale_color_continuous(palette = "Batlow", name = "Horsepower") |>
  scale_size(range = c(1, 8), name = "Displacement")
```

## Controlling a legend

`guides()` overrides a single legend without respelling its scale. Pass
`"none"` to hide it, or `guide_legend()` to reverse the key order or override
the title.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  guides(color = guide_legend(title = "Cylinders", reverse = TRUE))
```

Hiding a legend keeps the mapping; the marks stay coloured, only the guide
disappears:

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  guides(color = "none")
```

## Rich titles

Any scale `name` (and any `labs()` title) accepts `md()`, a small markdown
subset for bold, italic, superscript, subscript, and coloured spans. Handy for
units and formulae.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous(name = md("Power (hp m^2^)"))
```

Faceted plots add one more question: should panels share a scale or train their
own? That is the `resolve_scale()` lattice, covered in
**[Facets and composition](facets-and-composition.html)**. For stat-derived
aesthetics like `after_stat(count)`, see
**[Statistical marks](statistical-marks.html)**.
