---
title: "Scales and guides"
description: >
  How data values become colour, size, shape, and axis positions. Continuous,
  discrete, binned, manual, and gradient scales, trained domains, and the
  legends they produce.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("quill", quietly = TRUE)
)
library(quill)
```

A *scale* is the function that turns a data value into something you can see: a
colour, a marker size, a shape, or a position along an axis. In quill every
mapped encoding gets a scale automatically, trained from the data, so most
plots need no explicit scale at all. You reach for a `scale_*()` call only when
you want to override the default: change the palette, set a transform, fix the
limits, or rename the guide.

The key idea is *training*. When you map `color = hp`, quill scans the `hp`
values across every layer, works out the domain, and builds a continuous colour
scale with a legend. Add a `scale_*()` and you are declaring an override on top
of that trained default, not replacing the whole machine.

## Colour, continuous and discrete

Continuous data get a smooth colour ramp; discrete data get a categorical
palette. The `palette` argument takes a vector of colours or a single palette
name passed to `grDevices::hcl.colors()` (for example `"Batlow"`, `"Blues"`,
`"Set 2"`).

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3) |>
  scale_color_continuous(palette = "Viridis", name = "Horsepower")
```

For a two-colour ramp, `scale_color_gradient()` takes the endpoints directly.
For categories, `scale_color_manual()` maps levels to colours; naming the
values pins each level to a specific colour.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl), size = 3) |>
  scale_color_manual(values = c("4" = "#1b9e77", "6" = "#d95f02", "8" = "#7570b3"))
```

## Binned colour

`scale_fill_binned()` and `scale_color_binned()` cut a continuous aesthetic
into classes and give it a discrete legend, which is what you want for a
choropleth or a heatmap you read by band rather than by exact value. Choose the
classification `style` (`"quantile"`, `"equal"`, `"pretty"`, or any
`classInt::classIntervals()` style) and the number of classes `n`.

```{r}
grid <- expand.grid(x = 1:10, y = 1:10)
grid$z <- with(grid, x * y)
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_binned(style = "quantile", n = 5, palette = "Mako")
```

## Size, shape, and edge width

Non-colour aesthetics have scales too. `scale_size()` maps values linearly to a
marker-size `range` (in mm); `scale_shape()` cycles a set of shapes over the
levels of a discrete aesthetic; `scale_edge_width()` does for network edges
what `scale_size()` does for points.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = hp, color = factor(cyl)) |>
  scale_size(range = c(2, 9), name = "hp")
```

## Position scales

`scale_x_continuous()` and `scale_y_continuous()` control the axes. By default
they train from the data with a small expansion; override them to set `limits`,
apply a `trans` (`"log10"`, `"sqrt"`, `"reverse"`), or supply explicit `breaks`
and `labels`. Categorical axes use `scale_x_discrete()` and
`scale_y_discrete()`, whose `limits` set the order or subset of levels.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  scale_y_continuous(limits = c(10, 35), breaks = seq(10, 35, 5)) |>
  scale_x_continuous(trans = "log10", name = "Weight (log scale)")
```

## Guides come for free

Every scale that needs a legend produces one, and quill stacks multiple
legends automatically. Map two aesthetics and you get two guides without asking.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = disp) |>
  scale_color_continuous(palette = "Batlow", name = "Horsepower") |>
  scale_size(range = c(1, 8), name = "Displacement")
```

## Rich titles

Any scale `name` (and any `labs()` title) accepts `md()`, a small markdown
subset for bold, italic, superscript, subscript, and coloured spans. Handy for
units and formulae.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous(name = md("Power (hp m^2^)"))
```

Faceted plots add one more question: should panels share a scale or train their
own? That is the `resolve_scale()` lattice, covered in
**[Facets and composition](facets-and-composition.html)**. For stat-derived
aesthetics like `after_stat(count)`, see
**[Statistical marks](statistical-marks.html)**.
