---
title: "Scales and guides"
description: >
  How data values become colour, size, shape, opacity, line type, and axis
  positions. Continuous, discrete, binned, manual, gradient, date/time, and
  identity scales, limit shortcuts, trained domains, and controlling the legends
  they produce.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("quill", quietly = TRUE)
)
library(quill)
```

A *scale* is the function that turns a data value into something you can see: a
colour, a marker size, a shape, or a position along an axis. In quill every
mapped encoding gets a scale automatically, trained from the data, so most
plots need no explicit scale at all. You reach for a `scale_*()` call only when
you want to override the default: change the palette, set a transform, fix the
limits, or rename the guide.

The key idea is *training*. When you map `color = hp`, quill scans the `hp`
values across every layer, works out the domain, and builds a continuous colour
scale with a legend. Add a `scale_*()` and you are declaring an override on top
of that trained default, not replacing the whole machine.

## Colour, continuous and discrete

Continuous data get a smooth colour ramp; discrete data get a categorical
palette. The `palette` argument takes a vector of colours or a single palette
name passed to `grDevices::hcl.colors()` (for example `"Batlow"`, `"Blues"`,
`"Set 2"`).

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3) |>
  scale_color_continuous(palette = "Viridis", name = "Horsepower")
```

For a two-colour ramp, `scale_color_gradient()` takes the endpoints directly.
For categories, `scale_color_manual()` maps levels to colours; naming the
values pins each level to a specific colour.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl), size = 3) |>
  scale_color_manual(values = c("4" = "#1b9e77", "6" = "#d95f02", "8" = "#7570b3"))
```

## Binned colour

`scale_fill_binned()` and `scale_color_binned()` cut a continuous aesthetic
into classes and give it a discrete legend, which is what you want for a
choropleth or a heatmap you read by band rather than by exact value. Choose the
classification `style` (`"quantile"`, `"equal"`, `"pretty"`, or any
`classInt::classIntervals()` style) and the number of classes `n`.

```{r}
grid <- expand.grid(x = 1:10, y = 1:10)
grid$z <- with(grid, x * y)
vplot(grid) |>
  mark_tile(x = x, y = y, fill = z) |>
  scale_fill_binned(style = "quantile", n = 5, palette = "Mako")
```

## Size, shape, and edge width

Non-colour aesthetics have scales too. `scale_size()` maps values linearly to a
marker-size `range` (in mm); `scale_shape()` cycles a set of shapes over the
levels of a discrete aesthetic; `scale_edge_width()` does for network edges
what `scale_size()` does for points.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, size = hp, color = factor(cyl)) |>
  scale_size(range = c(2, 9), name = "hp")
```

## Opacity and line type

`alpha` and `linetype` are mapped aesthetics too. `scale_alpha()` maps a
continuous variable to opacity (its `range` defaults to `c(0.1, 1)`), a good way
to let density show through an overplotted cloud.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, alpha = hp, size = 3) |>
  scale_alpha(range = c(0.2, 1))
```

`scale_linetype()` maps a discrete variable to line types, cycling `"solid"`,
`"dashed"`, `"dotted"`, and so on. It applies to line-like marks
(`mark_line()`, `mark_step()`).

```{r}
series <- data.frame(
  x = rep(1:20, 3),
  y = c(cumsum(rnorm(20)), cumsum(rnorm(20)), cumsum(rnorm(20))),
  series = rep(c("a", "b", "c"), each = 20)
)
vplot(series) |>
  mark_line(x = x, y = y, linetype = series)
```

## Identity scales

Sometimes a column *already* holds the exact aesthetic values you want — actual
colour names, sizes in millimetres, shape names. An identity scale uses them
verbatim and draws no legend. There is a variant for each aesthetic:
`scale_color_identity()`, `scale_fill_identity()`, `scale_size_identity()`,
`scale_shape_identity()`, `scale_alpha_identity()`, `scale_linetype_identity()`.

```{r}
df <- data.frame(
  x = 1:5, y = 1:5,
  col = c("firebrick", "goldenrod", "forestgreen", "steelblue", "purple")
)
vplot(df) |>
  mark_point(x = x, y = y, color = col, size = 6) |>
  scale_color_identity()
```

## Position scales

`scale_x_continuous()` and `scale_y_continuous()` control the axes. By default
they train from the data with a small expansion; override them to set `limits`,
apply a `trans` (`"log10"`, `"sqrt"`, `"reverse"`), or supply explicit `breaks`
and `labels`. Categorical axes use `scale_x_discrete()` and
`scale_y_discrete()`, whose `limits` set the order or subset of levels.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  scale_y_continuous(limits = c(10, 35), breaks = seq(10, 35, 5)) |>
  scale_x_continuous(trans = "log10", name = "Weight (log scale)")
```

To set limits without spelling out a whole scale, use the shortcuts `xlim()`,
`ylim()`, and `lims()` (the last takes one named argument per aesthetic):

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  lims(x = c(0, 6), y = c(0, 40))
```

## Date and time axes

A `Date` or `POSIXct` column gets a date axis automatically. To control the
break interval or the label format, declare `scale_x_date()` (or
`scale_x_datetime()` / `scale_x_time()`): `date_breaks` takes an interval string
like `"6 months"`, and `date_labels` a `strftime()` format.

```{r}
econ <- data.frame(
  day = as.Date("2020-01-01") + 0:729,
  value = cumsum(rnorm(730))
)
vplot(econ) |>
  mark_line(x = day, y = value) |>
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y")
```

## Guides come for free

Every scale that needs a legend produces one, and quill stacks multiple
legends automatically. Map two aesthetics and you get two guides without asking.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = disp) |>
  scale_color_continuous(palette = "Batlow", name = "Horsepower") |>
  scale_size(range = c(1, 8), name = "Displacement")
```

## Controlling a legend

`guides()` overrides a single legend without respelling its scale. Pass
`"none"` to hide it, or `guide_legend()` to reverse the key order or override
the title.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  guides(color = guide_legend(title = "Cylinders", reverse = TRUE))
```

Hiding a legend keeps the mapping — the marks stay coloured, only the guide
disappears:

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  guides(color = "none")
```

## Rich titles

Any scale `name` (and any `labs()` title) accepts `md()`, a small markdown
subset for bold, italic, superscript, subscript, and coloured spans. Handy for
units and formulae.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous(name = md("Power (hp m^2^)"))
```

Faceted plots add one more question: should panels share a scale or train their
own? That is the `resolve_scale()` lattice, covered in
**[Facets and composition](facets-and-composition.html)**. For stat-derived
aesthetics like `after_stat(count)`, see
**[Statistical marks](statistical-marks.html)**.
