---
title: "Spatial and networks"
description: >
  Two data shapes that are not tidy rows of x and y: sf geometries drawn as
  maps with coord_sf, and igraph graphs laid out and drawn as node-link
  diagrams with vgraph.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
library(vellumplot)
have_sf <- requireNamespace("sf", quietly = TRUE)
have_net <- requireNamespace("igraph", quietly = TRUE) &&
  requireNamespace("graphlayouts", quietly = TRUE)
```

Most of vellumplot maps columns of a data frame to `x` and `y`. Two important data
shapes do not fit that mould: spatial geometries, where the coordinates live in
a geometry column and the aspect ratio matters, and graphs, where the positions
have to be computed by a layout algorithm first. vellumplot handles both with the
same grammar, through entry points suited to each data type.

## Maps from sf

`mark_sf()` draws the geometry column of an `sf` object. Points, lines, and
polygons each render appropriately, and there are no `x`/`y` encodings: the
coordinates come from the geometry. Other aesthetics map feature attributes as
usual, so `fill = AREA` gives you a choropleth. Pair it with `coord_sf()`,
which reprojects every layer to a common CRS before training and locks the map
aspect ratio so nothing is stretched.

```{r}
#| eval: !expr have_sf
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
vplot(nc) |>
  mark_sf(fill = BIR74) |>
  scale_fill_binned(style = "quantile", n = 6, palette = "Batlow") |>
  coord_sf()
```

Because `mark_sf()` is an ordinary layer, you can stack it: draw a base layer
of boundaries, then a second `mark_sf()` for highlighted features or point
locations on top, and every layer reprojects together.

## Networks from igraph

`vgraph()` starts a node-link diagram from an `igraph` graph. It runs a layout
(stress majorization by default, via `graphlayouts`) and produces a spec whose
node and edge tables already carry the `x`, `y`, `xend`, `yend`, and `name`
columns the graph marks need. Then `mark_edges()` and `mark_nodes()` draw it.
Draw order is fixed no matter how you pipe them: edges below nodes below
labels.

```{r}
#| eval: !expr have_net
g <- igraph::make_graph("Zachary")
g <- igraph::set_vertex_attr(
  g, "grp",
  value = as.factor(igraph::cluster_louvain(g)$membership)
)
g <- igraph::set_vertex_attr(g, "deg", value = igraph::degree(g))

vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.4) |>
  mark_nodes(size = deg, fill = grp) |>
  scale_size(range = c(2, 9))
```

The node and edge aesthetics are the same ones you already know. `mark_nodes()`
takes `size`, `shape`, `fill`, and `color`; `mark_edges()` takes `color`,
`linewidth`, `linetype`, and `alpha`. `mark_node_text()` adds vertex labels.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "stress") |>
  mark_edges(color = "grey70") |>
  mark_nodes(fill = grp, size = 6) |>
  mark_node_text(label = name, size = 8)
```

### Independent edge scales

Edge colour, opacity, line type, and width train on their **own** scales --
`scale_edge_color()`, `scale_edge_alpha()`, `scale_edge_linetype()`, and
`scale_edge_width()` -- separate from the node colour/alpha/linetype scales. So a
figure can map node fill to a discrete community *and* edge colour to a
continuous edge weight, and each gets its own legend instead of the two
collapsing into one. Map an edge attribute to `color` and it is trained as an
edge scale automatically; call `scale_edge_color()` to choose the palette.

```{r}
#| eval: !expr have_net
g <- igraph::set_edge_attr(g, "w", value = runif(igraph::ecount(g)))

vgraph(g, layout = "stress") |>
  mark_edges(color = w, linewidth = w) |>
  mark_nodes(fill = grp, size = deg) |>
  scale_edge_color(palette = "Grays") |>
  scale_edge_width(range = c(0.3, 2.5)) |>
  scale_size(range = c(2, 9))
```

### Directed edges, arrows, and edge labels

For directed graphs, `arrow = TRUE` draws a closed arrowhead at each edge's
target end; edges are capped at the node boundary so the head is never buried
under the marker. Pass a `vellum::vl_arrow()` for full control over the head
(`ends`, `type`, `length`, `angle`). `mark_edge_text()` labels the edges at
their midpoints, and `angle = "along"` rotates each label to follow its edge.

```{r}
#| eval: !expr have_net
el <- matrix(c(1, 2, 2, 3, 3, 1, 1, 4, 4, 2), ncol = 2, byrow = TRUE)
d <- igraph::graph_from_edgelist(el, directed = TRUE)
d <- igraph::set_edge_attr(d, "flow", value = c(3, 1, 4, 1, 5))

vgraph(d, layout = "stress") |>
  mark_edges(linewidth = flow, arrow = TRUE) |>
  mark_nodes(size = 8, fill = "steelblue") |>
  mark_edge_text(label = flow, size = 7) |>
  mark_node_text(label = name, size = 8, color = "white") |>
  scale_edge_width(range = c(0.4, 2.5))
```

### Readable labels

Labelling every vertex of a real network is unreadable, and labels that sit on
the markers or vanish into the edges are worse. `mark_node_text()` has four tools
for this:

- `top_n` / `by` label only the most important vertices (e.g. the highest-degree
  hubs) -- the common "label a handful, not all 34" move as a one-liner.
- `dist` pushes each label radially outward from the layout centre, so it clears
  its node marker.
- `repel = TRUE` nudges any labels that still overlap apart, ggrepel-style, with a
  thin leader line back to each vertex.
- `effects = list(shadow())` drops a shadow behind each label so it stays legible
  where it crosses an edge. `mark_edge_text()` takes the same `effects`.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = deg, fill = grp) |>
  mark_node_text(
    label = name,
    top_n = 8,
    by = deg,
    dist = 3,
    color = "white",
    effects = list(shadow(x = 0.4, y = -0.4))
  ) |>
  scale_size(range = c(2, 9))
```

### Edge routing

Edges are straight by default. `routing = "elbow"` draws orthogonal right-angle
steps instead -- still straight segments (curved edges are a deliberate
non-goal), stepping along whichever axis the endpoints are farther apart on, so a
top-down tree bends downward. It is the natural routing for `"tree"`,
`"sugiyama"`, and dendrogram layouts, and keeps node-boundary caps and
arrowheads.

```{r}
#| eval: !expr have_net
tr <- igraph::make_tree(15, children = 2, mode = "out")
vgraph(tr, layout = "tree") |>
  mark_edges(routing = "elbow", arrow = TRUE) |>
  mark_nodes(size = 5, fill = "steelblue")
```

For directed graphs, `gradient = TRUE` fades each edge from faint at its source
to opaque at its target -- a direction cue that reads without arrowheads (and
without the clutter they add on a dense graph).

```{r}
#| eval: !expr have_net
dg <- igraph::sample_gnp(15, 0.2, directed = TRUE)
vgraph(dg) |>
  mark_edges(gradient = TRUE) |>
  mark_nodes(size = 4, fill = "grey30")
```

### Augmenting and filtering

Vertex metrics are the analyst's choice, not something `vgraph()` computes behind
your back -- but `augment` attaches a set on request, so you can map or filter by
them without a manual igraph round-trip. `augment = TRUE` adds `degree` and
`components`; a character vector picks from `degree`, `betweenness`, `closeness`,
`eigen`, `coreness`, `components`, `community`, and the in/out-degree variants.

`filter_nodes` / `filter_edges` take data-masked predicates over the vertex / edge
attributes, and `k_core` keeps the k-core. Filtering happens **before** the
layout, so you see a clean layout of the subgraph rather than a full layout with
holes -- the honest way to tame a dense graph (the node-link idiom breaks down
past a link density of ~3).

```{r}
#| eval: !expr have_net
# attach degree + community, then plot just the 2-core, sized and coloured by them
vgraph(g, augment = c("degree", "community"), k_core = 2) |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = degree, fill = factor(community)) |>
  mark_node_text(label = name, top_n = 6, by = degree, dist = 3) |>
  scale_size(range = c(2, 9))
```

### Community hulls and node glyphs

`mark_node_hull()` shades groups of nodes with a convex hull drawn *behind* the
graph -- the standard way to show community structure. Group by a mapped `fill`;
`expand` grows each hull so it wraps the markers rather than clipping them.

```{r}
#| eval: !expr have_net
g <- igraph::set_vertex_attr(
  g, "comm",
  value = factor(igraph::membership(igraph::cluster_louvain(g)))
)
vgraph(g, layout = "stress") |>
  mark_node_hull(fill = comm, expand = 0.12) |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = deg, fill = comm) |>
  scale_size(range = c(2, 9))
```

`mark_node_pie()` replaces the node markers with pie (or donut, via `inner`)
glyphs whose wedges come from a set of compositional columns -- one wedge per
column, sized by that column's value at each vertex.

```{r}
#| eval: !expr have_net
set.seed(1)
for (nm in c("x1", "x2", "x3")) {
  g <- igraph::set_vertex_attr(g, nm, value = runif(igraph::vcount(g)))
}
vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.3) |>
  mark_node_pie(cols = c("x1", "x2", "x3"), size = 5)
```

### Choosing a layout

`layout` accepts a name (`"stress"`, `"sparse_stress"`, `"backbone"`, `"fr"`,
`"kk"`, `"circle"`, `"tree"`, `"sugiyama"`, and more), a supplied N-by-2
coordinate matrix, or a function that returns one. Stochastic layouts take a
`seed` so the figure is reproducible.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "circle") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(fill = grp, size = 5)
```

Both of these are still ordinary specs. They face the same scales, themes, and
composition tools as any other plot, and they render to a file with
`render_plot()`. For per-feature interactivity (tooltips and data keys on map
features or nodes), see the interactivity notes on `mark_sf()` and the mark
reference pages.
