---
title: "Spatial and networks"
description: >
  Two data shapes that are not tidy rows of x and y: sf geometries drawn as
  maps with coord_sf, and igraph graphs laid out and drawn as node-link
  diagrams with vgraph.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
library(vellumplot)
have_sf <- requireNamespace("sf", quietly = TRUE)
have_net <- requireNamespace("igraph", quietly = TRUE) &&
  requireNamespace("graphlayouts", quietly = TRUE)
have_bundle <- have_net && requireNamespace("edgebundle", quietly = TRUE)
```

Most of vellumplot maps columns of a data frame to `x` and `y`. Two important data
shapes do not fit that mould: spatial geometries, where the coordinates live in
a geometry column and the aspect ratio matters, and graphs, where the positions
have to be computed by a layout algorithm first. vellumplot handles both with the
same grammar, through entry points suited to each data type.

## Maps from sf

`mark_sf()` draws the geometry column of an `sf` object. Points, lines, and
polygons each render appropriately, and there are no `x`/`y` encodings: the
coordinates come from the geometry. Other aesthetics map feature attributes as
usual, so `fill = AREA` gives you a choropleth. Pair it with `coord_sf()`,
which reprojects every layer to a common CRS before training and locks the map
aspect ratio so nothing is stretched.

```{r}
#| eval: !expr have_sf
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
vplot(nc) |>
  mark_sf(fill = BIR74) |>
  scale_fill_binned(style = "quantile", n = 6, palette = "Batlow") |>
  coord_sf()
```

Because `mark_sf()` is an ordinary layer, you can stack it: draw a base layer
of boundaries, then a second `mark_sf()` for highlighted features or point
locations on top, and every layer reprojects together.

### Labelling features

`mark_sf_label()` names each feature at its **interior point**
(`sf::st_point_on_surface()`, guaranteed to fall inside the polygon rather than
in a bay or a hole). The label positions reproject through the same `coord_sf()`
CRS as the geometry, so they always land where the feature is drawn, and crowded
labels repel apart.

```{r}
vplot(nc) |>
  mark_sf(fill = AREA) |>
  mark_sf_label(label = NAME, size = 5) |>
  coord_sf()
```

### Merging regions

When several small features share a value — counties grouped into a region, or
tracts into a district — the borders *between* them are usually noise: you want
one clean outline per group, not a mesh of internal seams. `merge = TRUE`
dissolves each same-fill group into a single region with real boolean geometry
(`vellum::vl_path_op()`), so the shared edges vanish and each region is one crisp
path that stays exact in PDF (no double-stroked seams, no hairline gaps).

```{r}
#| eval: !expr have_sf
nc$region <- cut(sf::st_coordinates(sf::st_centroid(sf::st_geometry(nc)))[, 1],
  4,
  labels = c("west", "midwest", "mideast", "east")
)
vplot(nc) |>
  mark_sf(fill = region, merge = TRUE, color = "white") |>
  coord_sf()
```

Merging is for static choropleths; on an interactive layer it is ignored, since
dissolving the features would drop the per-feature keys a tooltip needs.

### Cartographic furniture

A map usually wants more than the polygons: a graticule to place it on the
globe, a scale bar to read distances, and a north arrow. `coord_sf(graticule =
TRUE)` draws meridians and parallels behind the marks — and because they are
generated in longitude/latitude and then reprojected, they curve correctly under
a projected CRS rather than being drawn as a naive straight grid.
`mark_scalebar()` and `mark_compass()` are fixed-position decorations pinned to a
panel corner (`position = "bottomleft"`, `"topright"`, and so on); the scale bar
reads the map's CRS to size itself, defaulting to a round distance near a quarter
of the panel width.

```{r}
#| eval: !expr have_sf
vplot(nc) |>
  mark_sf(fill = BIR74, color = "white", linewidth = 0.2) |>
  scale_fill_binned(style = "quantile", n = 6, palette = "Batlow") |>
  coord_sf(crs = 5070, graticule = TRUE) |>
  mark_scalebar(unit = "km") |>
  mark_compass()
```

## Networks from igraph

`vgraph()` starts a node-link diagram from an `igraph` graph. It runs a layout
(stress majorization by default, via `graphlayouts`) and produces a spec whose
node and edge tables already carry the `x`, `y`, `xend`, `yend`, and `name`
columns the graph marks need. Then `mark_edges()` and `mark_nodes()` draw it.
Draw order is fixed no matter how you pipe them: edges below nodes below
labels.

```{r}
#| eval: !expr have_net
g <- igraph::make_graph("Zachary")
g <- igraph::set_vertex_attr(
  g, "grp",
  value = as.factor(igraph::cluster_louvain(g)$membership)
)
g <- igraph::set_vertex_attr(g, "deg", value = igraph::degree(g))

vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.4) |>
  mark_nodes(size = deg, fill = grp) |>
  scale_size(range = c(0.5, 3))
```

The node and edge aesthetics are the same ones you already know. `mark_nodes()`
takes `size`, `shape`, `fill`, and `color`; `mark_edges()` takes `color`,
`linewidth`, `linetype`, and `alpha`. `mark_node_text()` adds vertex labels.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "stress") |>
  mark_edges(color = "grey70") |>
  mark_nodes(fill = grp, size = 2) |>
  mark_node_text(label = name, size = 8)
```

### Independent edge scales

Edge colour, opacity, line type, and width train on their **own** scales --
`scale_edge_color()`, `scale_edge_alpha()`, `scale_edge_linetype()`, and
`scale_edge_width()` -- separate from the node colour/alpha/linetype scales. So a
figure can map node fill to a discrete community *and* edge colour to a
continuous edge weight, and each gets its own legend instead of the two
collapsing into one. Map an edge attribute to `color` and it is trained as an
edge scale automatically; call `scale_edge_color()` to choose the palette.

```{r}
#| eval: !expr have_net
g <- igraph::set_edge_attr(g, "w", value = runif(igraph::ecount(g)))

vgraph(g, layout = "stress") |>
  mark_edges(color = w, linewidth = w) |>
  mark_nodes(fill = grp, size = deg) |>
  scale_edge_color(palette = "Grays") |>
  scale_edge_width(range = c(0.3, 2.5)) |>
  scale_size(range = c(0.5, 3))
```

### Directed edges, arrows, and edge labels

For directed graphs, `arrow = TRUE` draws a closed arrowhead at each edge's
target end; edges are capped at the node boundary so the head is never buried
under the marker. Pass a `vellum::vl_arrow()` for full control over the head
(`ends`, `type`, `length`, `angle`). `mark_edge_text()` labels the edges at
their midpoints, and `angle = "along"` rotates each label to follow its edge.

```{r}
#| eval: !expr have_net
el <- matrix(c(1, 2, 2, 3, 3, 1, 1, 4, 4, 2), ncol = 2, byrow = TRUE)
d <- igraph::graph_from_edgelist(el, directed = TRUE)
d <- igraph::set_edge_attr(d, "flow", value = c(3, 1, 4, 1, 5))

vgraph(d, layout = "stress") |>
  mark_edges(linewidth = flow, arrow = TRUE) |>
  mark_nodes(size = 2, fill = "steelblue") |>
  mark_edge_text(label = flow, size = 7) |>
  mark_node_text(label = name, size = 8, color = "white") |>
  scale_edge_width(range = c(0.4, 2.5))
```

### Readable labels

Labelling every vertex of a real network is unreadable, and labels that sit on
the markers or vanish into the edges are worse. `mark_node_text()` has four tools
for this:

- `top_n` / `by` label only the most important vertices (e.g. the highest-degree
  hubs) -- the common "label a handful, not all 34" move as a one-liner.
- `dist` pushes each label radially outward from the layout centre, so it clears
  its node marker.
- `repel = TRUE` nudges any labels that still overlap apart, ggrepel-style, with a
  thin leader line back to each vertex.
- `effects = list(shadow())` drops a shadow behind each label so it stays legible
  where it crosses an edge. `mark_edge_text()` takes the same `effects`.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = deg, fill = grp) |>
  mark_node_text(
    label = name,
    top_n = 8,
    by = deg,
    dist = 3,
    color = "white",
    effects = list(shadow(x = 0.4, y = -0.4))
  ) |>
  scale_size(range = c(0.5, 3))
```

### Edge routing

Edges are straight by default. `routing = "elbow"` draws orthogonal right-angle
steps instead -- still straight segments (curved edges are a deliberate
non-goal), stepping along whichever axis the endpoints are farther apart on, so a
top-down tree bends downward. It is the natural routing for `"tree"`,
`"sugiyama"`, and dendrogram layouts, and keeps node-boundary caps and
arrowheads.

```{r}
#| eval: !expr have_net
tr <- igraph::make_tree(15, children = 2, mode = "out")
vgraph(tr, layout = "tree") |>
  mark_edges(routing = "elbow", arrow = TRUE) |>
  mark_nodes(size = 2, fill = "steelblue")
```

For directed graphs, `gradient = TRUE` fades each edge from faint at its source
to opaque at its target -- a direction cue that reads without arrowheads (and
without the clutter they add on a dense graph).

```{r}
#| eval: !expr have_net
dg <- igraph::sample_gnp(15, 0.2, directed = TRUE)
vgraph(dg) |>
  mark_edges(gradient = TRUE) |>
  mark_nodes(size = 2, fill = "grey30")
```

### Edge bundling

On a dense graph, straight edges pile into an unreadable hairball. `mark_edge_bundle()`
routes them as bundled curves instead, so edges that run roughly together merge
into a few trunks and the backbone of the graph shows through. It is a drop-in
swap for `mark_edges()` -- the same edge aesthetics apply -- and delegates the
geometry to the [edgebundle](https://github.com/schochastics/edgebundle) package.
`type` picks the algorithm; bundled edges are faint by default so overlapping
trunks read as density.

```{r}
#| eval: !expr have_bundle
gb <- igraph::sample_gnp(60, 0.08)
vgraph(gb, layout = "stress") |>
  mark_edge_bundle(type = "hammer", color = "firebrick") |>
  mark_nodes(size = 1.5, fill = "grey20")
```

The other algorithms trade off speed against how aggressively they merge:
`"force"` (the default, force-directed) and `"path"` bend edges gently, `"stub"`
only tufts each endpoint, and `"mingle"` merges hierarchically. For a directed
graph, `type = "divided"` splits each trunk by direction. Pass algorithm-specific
tuning through `params`, e.g. `params = list(compatibility_threshold = 0.5)`.

### Flow maps

A *flow map* shows one source fanning out to many destinations, the branches
merging into trunks whose width grows with the combined flow -- the shape
Minard used for Napoleon's march and cartographers use for migration. Give
`mark_flow_map()` a one-to-many (star) graph laid out at fixed coordinates, name
the `root`, and map the per-edge flow to `weight`.

```{r}
#| eval: !expr have_bundle
set.seed(1)
dest <- 10
fg <- igraph::make_star(dest + 1, mode = "undirected")
igraph::V(fg)$name <- c("hub", paste0("d", seq_len(dest)))
igraph::E(fg)$weight <- sample(1:20, dest, replace = TRUE)
coords <- rbind(
  c(0, 0),
  cbind(runif(dest, 1, 6), runif(dest, -3, 3))
)

vgraph(fg, layout = coords) |>
  mark_flow_map(root = "hub", weight = weight) |>
  mark_nodes(size = 1.5, fill = "grey30")
```

`type = "spiral"` (the default) is the recommended layout: a planar,
angle-restricted spiral tree that needs only the edgebundle package.
`type = "steiner"` builds an approximate Steiner tree instead (it additionally
needs the \pkg{interp} package). The computed flow is mapped onto `width_range`,
so the widest trunk near the root stays legible however lopsided the weights are.

### Dendrograms

`vgraph()` also accepts a base `hclust`/`dendrogram`, coercing it to a tree that
carries each merge's height. The `"dendrogram"` layout places leaves on a line
and every merge at its height, and `elbow_at = "start"` turns the elbow into the
classic *bracket* (siblings share a bar at the parent's level):

```{r}
#| eval: !expr have_net
hc <- hclust(dist(USArrests))
vgraph(hc, layout = "dendrogram") |>
  mark_edges(routing = "elbow", elbow_at = "start", elbow_axis = "v")
```

Add labels with `mark_text()`, which controls `angle` and justification -- for a
top-down tree the leaf labels read vertically, hanging below their leaves (merge
nodes carry a blank label, so only the leaves show):

```{r}
#| eval: !expr have_net
vgraph(hc, layout = "dendrogram") |>
  mark_edges(routing = "elbow", elbow_at = "start", elbow_axis = "v") |>
  mark_text(
    x = x, y = y, label = label, size = 1.6,
    angle = 90, hjust = "right", nudge_y = -1.5
  )
```

`vdendrogram()` is the one-line preset for all of that -- bracket edges and
placed leaf labels, `direction` to orient it, and `k` to cut the tree and colour
the clusters (branches above the cut stay neutral):

```{r}
#| eval: !expr have_net
vdendrogram(hc, k = 3)
```

For unrooted trees (phylogeny-style), `layout = "unrooted"` uses graphlayouts'
`layout_as_tree_unrooted()` -- `mode` picks `"equalangle"`, `"equaldaylight"`, or
`"stress"`:

```{r}
#| eval: !expr have_net && packageVersion("graphlayouts") >= "1.2.5"
vgraph(igraph::make_tree(31, 3, "undirected"), layout = "unrooted", mode = "equaldaylight") |>
  mark_edges(alpha = 0.6) |>
  mark_nodes(size = 1.5, fill = "grey30")
```

### Augmenting and filtering

Vertex metrics are the analyst's choice, not something `vgraph()` computes behind
your back -- but `augment` attaches a set on request, so you can map or filter by
them without a manual igraph round-trip. `augment = TRUE` adds `degree` and
`components`; a character vector picks from `degree`, `betweenness`, `closeness`,
`eigen`, `coreness`, `components`, `community`, and the in/out-degree variants.

`filter_nodes` / `filter_edges` take data-masked predicates over the vertex / edge
attributes, and `k_core` keeps the k-core. Filtering happens **before** the
layout, so you see a clean layout of the subgraph rather than a full layout with
holes -- the honest way to tame a dense graph (the node-link idiom breaks down
past a link density of ~3).

```{r}
#| eval: !expr have_net
# attach degree + community, then plot just the 2-core, sized and coloured by them
vgraph(g, augment = c("degree", "community"), k_core = 2) |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = degree, fill = factor(community)) |>
  mark_node_text(label = name, top_n = 6, by = degree, dist = 3) |>
  scale_size(range = c(0.5, 3))
```

### Community hulls and node glyphs

`mark_node_hull()` shades groups of nodes with a convex hull drawn *behind* the
graph -- the standard way to show community structure. Group by a mapped `fill`;
`expand` grows each hull so it wraps the markers rather than clipping them.

```{r}
#| eval: !expr have_net
g <- igraph::set_vertex_attr(
  g, "comm",
  value = factor(igraph::membership(igraph::cluster_louvain(g)))
)
vgraph(g, layout = "stress") |>
  mark_node_hull(fill = comm, expand = 0.12) |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = deg, fill = comm) |>
  scale_size(range = c(0.5, 3))
```

`mark_node_pie()` replaces the node markers with pie (or donut, via `inner`)
glyphs whose wedges come from a set of compositional columns -- one wedge per
column, sized by that column's value at each vertex.

```{r}
#| eval: !expr have_net
set.seed(1)
for (nm in c("x1", "x2", "x3")) {
  g <- igraph::set_vertex_attr(g, nm, value = runif(igraph::vcount(g)))
}
vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.3) |>
  mark_node_pie(cols = c("x1", "x2", "x3"), size = 5)
```

### Interactive neighbour highlighting

`select_neighbours()` is a network-aware selection preset: in an interactive host
(`vellumwidget::as_widget()`), pointing at a node spotlights **its neighbourhood**
— the node, its incident edges, and its adjacent nodes — and dims the rest;
pointing at an edge highlights its two endpoints. `degree = 2` reaches
neighbours-of-neighbours. Like every selection it is inert on a static render, so
the code below draws the same picture on the page but comes alive under
`as_widget()`.

```{r}
#| eval: false
vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(size = deg, fill = grp) |>
  mark_node_text(label = name, top_n = 8, by = deg) |>
  select_neighbours(on = "hover") |>
  scale_size(range = c(0.5, 3))
```

Declaring the selection also keys every node (by its vertex name) and edge, so
node/edge tooltips, click-select, and pan/zoom work in the widget too.

### Choosing a layout

`layout` accepts a name (`"stress"`, `"sparse_stress"`, `"backbone"`, `"fr"`,
`"kk"`, `"circle"`, `"tree"`, `"sugiyama"`, `"dendrogram"`, `"unrooted"`, and
more), a supplied N-by-2 coordinate matrix, or a function that returns one.
Stochastic layouts take a `seed` so the figure is reproducible.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "circle") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(fill = grp, size = 2)
```

Both of these are still ordinary specs. They face the same scales, themes, and
composition tools as any other plot, and they render to a file with
`render_plot()`. For per-feature interactivity (tooltips and data keys on map
features or nodes), see the interactivity notes on `mark_sf()` and the mark
reference pages.
