---
title: "Spatial and networks"
description: >
  Two data shapes that are not tidy rows of x and y: sf geometries drawn as
  maps with coord_sf, and igraph graphs laid out and drawn as node-link
  diagrams with vgraph.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%"
)
library(vellumplot)
have_sf <- requireNamespace("sf", quietly = TRUE)
have_net <- requireNamespace("igraph", quietly = TRUE) &&
  requireNamespace("graphlayouts", quietly = TRUE)
```

Most of vellumplot maps columns of a data frame to `x` and `y`. Two important data
shapes do not fit that mould: spatial geometries, where the coordinates live in
a geometry column and the aspect ratio matters, and graphs, where the positions
have to be computed by a layout algorithm first. vellumplot handles both with the
same grammar, through entry points suited to each data type.

## Maps from sf

`mark_sf()` draws the geometry column of an `sf` object. Points, lines, and
polygons each render appropriately, and there are no `x`/`y` encodings: the
coordinates come from the geometry. Other aesthetics map feature attributes as
usual, so `fill = AREA` gives you a choropleth. Pair it with `coord_sf()`,
which reprojects every layer to a common CRS before training and locks the map
aspect ratio so nothing is stretched.

```{r}
#| eval: !expr have_sf
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
vplot(nc) |>
  mark_sf(fill = BIR74) |>
  scale_fill_binned(style = "quantile", n = 6, palette = "Batlow") |>
  coord_sf()
```

Because `mark_sf()` is an ordinary layer, you can stack it: draw a base layer
of boundaries, then a second `mark_sf()` for highlighted features or point
locations on top, and every layer reprojects together.

## Networks from igraph

`vgraph()` starts a node-link diagram from an `igraph` graph. It runs a layout
(stress majorization by default, via `graphlayouts`) and produces a spec whose
node and edge tables already carry the `x`, `y`, `xend`, `yend`, and `name`
columns the graph marks need. Then `mark_edges()` and `mark_nodes()` draw it.
Draw order is fixed no matter how you pipe them: edges below nodes below
labels.

```{r}
#| eval: !expr have_net
g <- igraph::make_graph("Zachary")
g <- igraph::set_vertex_attr(
  g, "grp",
  value = as.factor(igraph::cluster_louvain(g)$membership)
)
g <- igraph::set_vertex_attr(g, "deg", value = igraph::degree(g))

vgraph(g, layout = "stress") |>
  mark_edges(alpha = 0.4) |>
  mark_nodes(size = deg, fill = grp) |>
  scale_size(range = c(2, 9))
```

The node and edge aesthetics are the same ones you already know. `mark_nodes()`
takes `size`, `shape`, `fill`, and `color`; `mark_edges()` takes `color`,
`linewidth` (scaled by `scale_edge_width()`), and `alpha`, plus `arrow = TRUE`
for directed graphs. `mark_node_text()` adds vertex labels.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "stress") |>
  mark_edges(color = "grey70") |>
  mark_nodes(fill = grp, size = 6) |>
  mark_node_text(label = name, size = 8)
```

### Choosing a layout

`layout` accepts a name (`"stress"`, `"sparse_stress"`, `"backbone"`, `"fr"`,
`"kk"`, `"circle"`, `"tree"`, `"sugiyama"`, and more), a supplied N-by-2
coordinate matrix, or a function that returns one. Stochastic layouts take a
`seed` so the figure is reproducible.

```{r}
#| eval: !expr have_net
vgraph(g, layout = "circle") |>
  mark_edges(alpha = 0.3) |>
  mark_nodes(fill = grp, size = 5)
```

Both of these are still ordinary specs. They face the same scales, themes, and
composition tools as any other plot, and they render to a file with
`render_plot()`. For per-feature interactivity (tooltips and data keys on map
features or nodes), see the interactivity notes on `mark_sf()` and the mark
reference pages.
