---
title: "Specs, agents, and interoperability"
description: >
  A vellumplot plot is a serializable document. This article shows the spec
  round-trip, how an LLM / MCP agent generates and validates plots, the
  Vega-Lite bridge (with a coverage table), and data-traceable provenance.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE) &&
    requireNamespace("jsonlite", quietly = TRUE)
)
library(vellumplot)
```

Because a `PlotSpec` is plain, inspectable data behind a compiler (see
[the compiler article](the-compiler.html)), a plot can be turned into a portable
*document* — serialized, generated by a model, translated to another engine, and
traced back to its data. This article covers the four capabilities that build on
that.

## The spec round-trip

`as_spec()` turns a plot into a nested list; `from_spec()` rebuilds it. The JSON
form is the wire format.

```{r}
p <- vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = factor(cyl)) |>
  labs(title = "Fuel economy")

spec <- as_spec(p)
str(spec$layers[[1]], max.level = 2)

json <- spec_to_json(p)
identical(as_spec(spec_from_json(json))$layers, as_spec(p)$layers)
```

The serializer covers the **encoding-level grammar** exactly. Anything a portable
document cannot carry — custom transform *functions*, paint/pattern fills,
hand-drawn `sketch()` geometry, secondary axes, `sf` CRS objects, per-layer data
— is **refused** with a classed `vellumplot_unserializable` error naming the
slot, never silently dropped. (Theme is the one exception: the preset name and
scalar settings round-trip; custom element overrides drop with a warning, since
styling is orthogonal to the data spec.) Data is inlined when small and stored by
reference (a content hash + column schema) otherwise.

## Generating plots from an agent

The serializable spec makes vellumplot a safe target for LLM / agent generation:
the agent emits *data* (a spec) that you validate and render with your own
trusted code — never `eval()`-d R. Three pieces support the loop.

`spec_fields()` grounds the model in the real columns and inferred types:

```{r}
spec_fields(mtcars)
```

`vplot_from_spec()` validates a generated spec against the data and returns
structured diagnostics — the machine-readable repair loop — instead of a
traceback:

```{r}
bad <- spec
bad$layers[[1]]$encoding$x$field <- "weight" # a hallucinated column
d <- spec_diagnose(bad, data = mtcars)
d$ok
d$diagnostics[[1]]
```

`mcp_serve()` runs a pure-R [Model Context Protocol](https://modelcontextprotocol.io)
server exposing `get_schema`, `list_fields`, and `render_spec` tools. Point an
agent at the bundled launcher:

```sh
Rscript -e 'vellumplot::mcp_serve()'
```

`vplot_ask()` is a model-agnostic convenience: give it a `responder` function
(wrapping any model or MCP client) and it assembles the grounding payload and
validates the result.

## Vega-Lite interoperability

`spec_to_vegalite()` and `spec_from_vegalite()` translate a plot to and from a
[Vega-Lite](https://vega.github.io/vega-lite/) specification.

```{r}
vl <- spec_to_vegalite(
  vplot(mtcars) |> mark_point(x = wt, y = mpg, color = cyl)
)
str(vl, max.level = 2)
```

### Coverage

The bridge covers a documented subset; anything it cannot map is reported (a
warning), never silently diverged.

| vellumplot | Vega-Lite | mapped |
|---|---|---|
| `mark_point/line/area/bar/rule/text/boxplot/rug/errorbar` | `point/line/area/bar/rule/text/boxplot/tick/errorbar` | ✅ both ways |
| `mark_tile`, `mark_pie`/`donut`, `mark_step` | `rect`, `arc`, `line` (`step-after`) | ✅ export |
| encodings `x/y/color/fill/size/shape/alpha/label`, `xmin/xmax/ymin/ymax` | `x/y/color/size/shape/opacity/text`, `x2/y2` | ✅ |
| `scale_*` domain, `trans` (log/sqrt), categorical range, `name` | `encoding.scale.{domain,type,range}`, `.title` | ✅ |
| `mark_histogram` (`stat="bin"`) | `bin` + `aggregate:"count"` | ✅ both ways |
| `facet_wrap` / `facet_grid` | `facet` / `row` + `column` | ✅ |
| inline data, plot title | `data.values`, `title` | ✅ |
| layer effects, patterns, sketch | — | ⚠ dropped (reported) |
| polar / flipped coordinates | — | ⚠ dropped (reported) |
| statistical stats (`smooth`, `density`, …) | — | ⚠ dropped (reported) |

## Provenance: tracing a plot back to its data

Every drawn element keeps a stable id tying it to the data rows that produced it.
`provenance_join()` surfaces that alongside the element's device-pixel geometry:

```{r}
pj <- provenance_join(p)
pj[, c("id", "mark", "n_rows")]
pj$rows[[1]] # the source-data rows behind the first element
```

A figure can also carry a fingerprint of its data, so it verifies against the
data it was drawn from:

```{r}
svg <- plot_svg(p, manifest = TRUE)
plot_verify(svg, mtcars)$ok # TRUE
plot_verify(svg, mtcars[1:5, ])$ok # FALSE — different data
```

`provenance_payload()` exposes the same element → rows mapping as a payload a
host (such as vellumwidget) can use for click-to-source interaction.
