---
title: "Statistical marks"
description: >
  Marks that transform data before drawing: histograms, densities, per-group
  summaries, linear smooths, ECDFs, Q-Q plots, rugs, violins, ridgelines, and
  dot plots, plus after_stat for reaching the computed variables and
  mark_datashade for millions of points.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("quill", quietly = TRUE)
)
library(quill)
```

Most marks draw the data as given. A statistical mark computes something from
the data first, then draws the result. quill runs that transform as a stage of
the compiler (see **[The compiler](the-compiler.html)**), so a histogram is not
you binning the data by hand; it is a mark that bins as part of being drawn.

## Histograms and densities

`mark_histogram()` bins a continuous `x` and draws the per-bin counts as bars.
Control the resolution with `bins`.

```{r}
vplot(faithful) |>
  mark_histogram(x = waiting, bins = 25)
```

`mark_density()` draws a smooth kernel-density estimate of `x` as a filled
curve; `adjust` scales the bandwidth. Densities are on a comparable vertical
scale, so mapping `fill` to a group and overlaying works well.

```{r}
vplot(iris) |>
  mark_density(x = Sepal.Length, fill = Species, alpha = 0.4)
```

## Per-group summaries

`mark_summary()` aggregates `y` within each `x` category using `fun` (the mean
by default) and draws the result. It is the quick way to add group means over
raw data.

```{r}
vplot(mtcars) |>
  mark_point(x = factor(cyl), y = mpg, color = "grey60") |>
  mark_summary(x = factor(cyl), y = mpg, fun = median)
```

## Smooths

`mark_smooth()` fits a model of `y` on `x` (`method = "lm"`) and draws the
fitted line with a confidence ribbon when `se = TRUE`. It layers naturally over
the raw points.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg, se = TRUE, level = 0.95)
```

## Distributions

Several marks summarise how a single variable is distributed. `mark_ecdf()`
draws the empirical cumulative distribution of `x` as a step, which is a
scale-free way to compare groups without choosing a bandwidth.

```{r}
vplot(iris) |>
  mark_ecdf(x = Sepal.Length, color = Species)
```

`mark_qq()` plots the sorted `sample` against the quantiles of a reference
distribution (normal by default); `mark_qq_line()` adds the reference line
through the quartiles. Points on the line mean the sample matches the reference.

```{r}
vplot(mtcars) |>
  mark_qq(sample = mpg) |>
  mark_qq_line(sample = mpg)
```

`mark_rug()` adds marginal ticks at each observation, a compact companion to a
scatter or density. `sides` picks the edges (`"b"`, `"l"`, `"t"`, `"r"`).

```{r}
vplot(faithful) |>
  mark_point(x = waiting, y = eruptions) |>
  mark_rug()
```

## Density shapes

`mark_violin()` draws a mirrored kernel density of `y` for each categorical
`x` — a boxplot's silhouette, showing the full shape of each group.

```{r}
vplot(iris) |>
  mark_violin(x = Species, y = Sepal.Length, fill = Species)
```

`mark_ridgeline()` turns that on its side: a density of `x` per categorical `y`,
with the ridges overlapping so many groups fit in little vertical space. Use the
`scale` argument to tune how much they overlap.

```{r}
vplot(iris) |>
  mark_ridgeline(x = Sepal.Length, y = Species, fill = Species)
```

`mark_dotplot()` bins `x` and stacks one dot per observation, so the height of
each stack is a count you can read dot by dot.

```{r}
vplot(faithful) |>
  mark_dotplot(x = waiting)
```

## 2-D density contours

`mark_contour()` estimates the 2-D density of a point cloud and draws its
iso-density contour lines; `mark_contour_filled()` fills the bands between them.
Both are coloured by level automatically. They read best over the points they
summarise:

```{r}
vplot(faithful) |>
  mark_point(x = eruptions, y = waiting, color = "grey70") |>
  mark_contour(x = eruptions, y = waiting)
```

```{r}
vplot(faithful) |>
  mark_contour_filled(x = eruptions, y = waiting)
```

The density estimate uses `MASS::kde2d()`; tune the levels with `bins`,
`binwidth`, or explicit `breaks`. To contour a *surface* you already have (a `z`
value on a regular `x`/`y` grid) rather than a point density, map `z`:

```{r}
grid <- expand.grid(x = seq(-3, 3, 0.1), y = seq(-3, 3, 0.1))
grid$z <- with(grid, dnorm(x) * dnorm(y))
vplot(grid) |> mark_contour(x = x, y = y, z = z)
```

Contour tracing needs the `isoband` package (and `MASS` for the density estimate).

## Reaching computed variables with after_stat

A statistical mark produces new variables that were not in your data: a
histogram computes a `count` and a `density`, for example. `after_stat()` lets
an encoding refer to one of those computed variables instead of a data column.
The classic use is a density-scaled histogram.

```{r}
vplot(faithful) |>
  mark_histogram(x = waiting, bins = 25, fill = after_stat(density))
```

Because the aesthetic is now driven by a computed value, its scale trains on
that value like any other, and you get the matching legend.

## Millions of points with mark_datashade

When there are too many points to draw one marker each (overplotted, up to
millions), individual markers stop being informative and start being slow.
`mark_datashade()` bins the points into a canvas-sized grid in a single pass and
colours each cell by density, drawing one raster. Its cost is decoupled from
the number of points.

```{r}
n <- 5e5
big <- data.frame(
  x = rnorm(n),
  y = rnorm(n) + rep(c(-1, 1), each = n / 2)
)
vplot(big) |>
  mark_datashade(x = y, y = x, how = "eq_hist")
```

Here `how = "eq_hist"` uses histogram equalisation so both dense and sparse
regions stay visible; the grid resolution is set by `width` and `height`.
Per-point aesthetics do not apply, because cell colour encodes density rather
than any single point.

`mark_point()` also has an `auto = TRUE` switch that falls back to a datashaded
raster automatically when a layer has very many rows, so you can keep writing
`mark_point()` and let quill choose the representation.
