---
title: "Get started"
description: >
  From a data frame to a rendered plot: build a spec, layer marks, train
  scales, and render it into a vellum scene.
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  out.width = "100%",
  eval = requireNamespace("vellumplot", quietly = TRUE)
)
```

vellumplot is a declarative, pipe-first grammar of graphics built on the
[vellum](https://r-vellum.github.io/vellum/) backend. You describe a plot
as an inspectable, serializable *spec*; nothing is drawn until the spec is
compiled into a vellum scene and rendered.

## Install

vellumplot compiles a Rust crate (inside `vellum`), so you need a Rust toolchain
(`cargo`/`rustc`) on your machine alongside R. With that in place:

```{r}
#| eval: false
# install.packages("pak")
pak::pak("r-vellum/vellumplot")
```

## A first plot

Start with `vplot()` on a data frame, then add a mark. Encodings (`x`, `y`,
`color`, …) are tidy-eval expressions evaluated against the data.

```{r}
library(vellumplot)

vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```

Printing a spec draws it into the plots pane (and embeds in a knitr/Quarto
chunk), like ggplot2.

## Layering marks

Add more marks to the same panel; scales train across every layer.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg)
```

## Faceting

Split into a grid of panels with shared or free scales.

```{r}
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  facet_wrap(~cyl)
```

## The spec is just data

Nothing is drawn until the spec is compiled. `summary()` shows its structure
without rendering.

```{r}
summary(vplot(mtcars) |> mark_point(x = wt, y = mpg, color = hp))
```

## Rendering to a file

`render_plot()` writes the compiled scene; the format follows the file
extension.

```{r}
#| eval: false
p <- vplot(mtcars) |> mark_point(x = wt, y = mpg)
render_plot(p, "cars.png")
```

## Where to next

- **[Reference](../reference/index.html)**: every mark, scale, coord, facet,
  theme, and effect.
- **[vellum](https://r-vellum.github.io/vellum/)**: the graphics backend
  vellumplot compiles into.
- **[vellumwidget](https://r-vellum.github.io/vellumwidget/)**: turn a scene into an
  interactive HTML widget.
- **[vellumverse](https://r-vellum.github.io/vellumverse/)**: install and
  attach the whole ecosystem in one step.
